---
title: "Bocconi University - 20236 Time Series Analysis"
subtitle: "**Group 27 | Anna-Bella Dreyfus, Emanuele Artoni, Artūrs Jānis Ņikitins, Madelief van Weerdenburg**"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: yes
    extra_dependencies: ["float"]
fontsize: 10pt
geometry: margin=2cm
header-includes:
  - \setlength{\textfloatsep}{6pt}
  - \setlength{\intextsep}{6pt}
---

```{r setup, include=FALSE}
rm(list = ls())
library(forecast)
library(tidyverse)
library(gridExtra)
library(knitr)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center", fig.pos = "H", out.extra = "")

set_theme(theme_classic())
update_theme(plot.title = element_text(hjust = 0.5), strip.background = element_blank())

plot_ts_gg <- function(
    x,
    title = NULL,
    xlab = NULL,
    ylab = NULL,
    aes = NULL,
    facet_ncol = 1,
    wrap_width = 45
) {
  if (!inherits(x, "ts")) {
    stop("Input must be ts.")
  }

  df <- tibble(
    Time = as.numeric(time(x)),
    Value = as.numeric(x)
  )

  p <- ggplot(df, aes(Time, Value, !!!aes)) +
    geom_line() +
    labs(title = title, x = NULL, y = ylab)

  p
}
```

# Exploratory Analysis

```{r, fig.height=3, fig.width=6, fig.show = "hold", out.width = "60%", fig.cap = "\\label{fig:main}Monthly Northern Hemisphere temperature anomalies (GISTEMP)."}
df <- read.csv("gistemp.csv", skip = 1, na.strings = "***")
#head(df)
monthly_matrix <- as.matrix(df[, 2:13])
monthly_vector <- as.numeric(t(monthly_matrix))

gistemp_ts <- ts(monthly_vector, start = c(1880, 1), frequency = 12)

plot_ts_gg(
  gistemp_ts,
  xlab = "Year",
  ylab = "Temperature anomalies (°C)"
  ) +
  geom_hline(yintercept = 0)


```

The time series presented in Figure$~$\ref{fig:main} shows a long-term upward trend in temperature anomalies over the sample period. Values are mostly negative before the mid-20th century and increase gradually over time. From approximately the 1980s onward, the rise becomes more pronounced, suggesting a possible acceleration in warming. Short-term fluctuations remain visible around this trend.

The marginal variances appear relatively stable, as the dispersion does not increase with the level of the series. There is no strong visual evidence of heteroskedasticity. The variance can therefore be considered approximately constant over time. This observation supports the use of an additive decomposition instead of a multiplicative one.

## Task 1

```{r task1_decompo_prep, message=FALSE, warning=FALSE}
# Ensuring positive values, necessary for classical multiplicative decomposition
shift <- 0
if (min(gistemp_ts, na.rm = TRUE) <= 0) {
  shift <- abs(min(gistemp_ts, na.rm = TRUE)) + 0.01
}
gistemp_pos <- gistemp_ts + shift
```

```{r task1_decompo, fig.height=4, fig.width=4, fig.show = "hold", out.width = "32%", fig.cap = "\\label{fig:decomp}Comparison of time-series decompositions of Northern Hemisphere temperature anomalies. For the multiplicative decomposition, the time series is shifted upwards by its most negative value."}
dec_add  <- decompose(gistemp_ts,  type = "additive")
dec_mult <- decompose(gistemp_pos, type = "multiplicative")
stl_fit  <- stl(gistemp_ts, s.window = "periodic")

# Assuming same seasonality for all decompositions
const_seasonality <- frequency(gistemp_ts)

# Conversion of the decompose() object for a cleaner output
df_decomp <- function(dec_obj) {
  data.frame(
    t = as.numeric(time(dec_obj$x)),
    observed = as.numeric(dec_obj$x),
    trend = as.numeric(dec_obj$trend),
    seasonal = as.numeric(dec_obj$seasonal),
    residual = as.numeric(dec_obj$random)
  )
}

df_stl <- function(stl_obj, y) {
  t <- as.numeric(time(stl_obj$time.series[, 1])) #extracting time index
  data.frame(
    t = t,
    observed = as.numeric(y),
    trend = as.numeric(stl_obj$time.series[, "trend"]),
    seasonal = as.numeric(stl_obj$time.series[, "seasonal"]),
    residual = as.numeric(stl_obj$time.series[, "remainder"])
  )
}

df_hw <- function(hw_obj, y, pad = 2) {
  t <- as.numeric(time(y))
  tibble(
    t = t,
    observed = as.numeric(y),
    fitted = c(rep(NA, 2), as.numeric(fitted(hw_obj)[, "xhat"])),
    trend = c(rep(NA, 2), as.numeric(fitted(hw_obj)[, "level"]) + as.numeric(fitted(hw_obj)[, "trend"])),
    residual = observed - fitted
  )
}

# Formatting as a panel using subplots
panel <- function(df, main_title, add_hline = TRUE) {
  p_observed <- ggplot(df, aes(t, observed)) + geom_line(size = 0.3) +
    labs(title = "Observed", x = NULL, y = NULL)
  if (add_hline) {
    p_observed <- p_observed + geom_hline(yintercept = 0)
  }
  p_trend <- ggplot(df, aes(t, trend)) + geom_line(size = 0.3) +
    labs(title = "Trend", x = NULL, y = NULL)

  seasonal_seq <- seq(from = 1, to = const_seasonality)

  p_seasonal <- df %>%
    slice_head(n = const_seasonality) %>%
    mutate(t = seasonal_seq) %>%
    ggplot(aes(t, seasonal)) + geom_line(size = 0.3) +
    scale_x_continuous(breaks = seasonal_seq) +
    labs(title = "Seasonal", x = NULL, y = NULL)


  p_residual <- ggplot(df, aes(t, residual)) + geom_line(size = 0.3) +
    labs(title = "Residual", x = NULL, y = NULL)

  # Choosing ncol = 2
  gridExtra::grid.arrange(
    p_observed, p_trend,
    p_seasonal, p_residual,
    ncol = 2,
    top = grid::textGrob(main_title, gp = grid::gpar(fontface = "bold", fontsize = 14))
  )
}

# Using grid to display panels

grid::grid.draw(panel(df_decomp(dec_add),  "Additive decomposition (GISTEMP)"))
grid::grid.draw(panel(df_decomp(dec_mult), "Multiplicative decomposition", add_hline = FALSE))
grid::grid.draw(panel(df_stl(stl_fit, gistemp_ts), "STL decomposition (periodic seasonality)"))

```

Given $m_t$, the deseasonalized mean level at time t, and $S_t$, the seasonal effect at time t, additive and multiplicative decompositions imply the following models, respectively:

$$
Y_t = m_t + S_t + \epsilon_t
$$
$$
Y_t = m_t S_t \epsilon_t
$$

In Figure$~$\ref{fig:decomp}, we can initially note that STL decomposition, which presupposes an additive structure, produces visually similar results to those of classical additive decomposition. As first evidenced by the exploratory analysis of the time series, the apparent lack of heteroskedasticity makes additive decomposition preferable. In particular, multiplicative decomposition induces the residuals' variance to be higher when the trend is lower, and vice-versa. Therefore, using an additive decomposition is more consistent with the variance structure of the original data, and thus the better choice.

Figure$~$\ref{fig:decomp} clearly indicates a long-term warming pattern in the GISTEMP data. After relatively moderate fluctuations in the late 19th and early 20th centuries, temperatures begin to increase gradually. A temporary stabilization is visible around the mid-20th century, with a relatively flatter slope between 1940 and 1970 suggesting a slowdown in warming. However, from the late 1970s onward, the trend exhibits a much steeper slope, indicating a strong and persistent acceleration in warming.

As, in the multiplicative decomposition, the seasonal component is expressed as a multiplicative factor, the plot is centered around 1, whereas in the additive (and STL) decomposition it is centered around 0.

While the overall temperature level has been steadily increasing due to the trend component, seasonality is characterized by small amplitude and stable magnitude, which does not appear to scale with the temperature level. For this reason, we have considered it appropriate to specify a periodic window width in our STL decomposition; had we determined otherwise, specifying a numerical value to this setting would have allowed the seasonal pattern to slowly evolve over time.

To better understand the intra-annual structure of the seasonal component, we now examine the annual seasonal profile by month. The monthly profiles confirm a clear seasonal cycle. For instance, late summer months such as August tend to exhibit negative seasonal effects, whereas autumn months such as October often contribute positively to the seasonal component.

The validity of our choice to impose an additive structure is further confirmed by an analysis of the residuals $\epsilon_t$, which appear to be devoid of any structure, centered at zero and with constant variance. While occasional extreme spikes are visible, they remain isolated and do not suggest structural misspecification.

## Task 2

### SES or Holt-Winters

```{r, fig.height = 3, fig.width = 6, fig.show = "hold", out.width = "60%", fig.cap = "\\label{fig:hw}Deseasonalized Northern Hemisphere temperature anomalies (1950–2010) and Holt–Winters one-step ahead forceasts."}
seasonal_component <- stl_fit$time.series[, "seasonal"]
deseasonalized <- gistemp_ts - seasonal_component
train <- window(deseasonalized, start = c(1950, 1), end = c(2010, 12))

## HW with trend (gamma = FALSE)
fit_holtwinters <- HoltWinters(train, gamma = FALSE, optim.start = c(alpha = 0.8, beta = 0.9))

# Parameters extraction
alpha_holtwinters <- fit_holtwinters$alpha
beta_holtwinters  <- fit_holtwinters$beta

#alpha_holtwinters
#beta_holtwinters

plot_ts_gg(train,
  ylab = "Temperature anomaly (°C)",
  aes = list(color = "Observed"),
) +
  geom_hline(yintercept = 0) +
  geom_line(data = df_hw(fit_holtwinters, train), aes(x = t, y = fitted, color = "One-step-ahead forecast")) +
  scale_color_manual(values = c("black", "steelblue")) +
  theme(legend.position = "inside", legend.position.inside = c(0.3, 0.9)) +
  labs(color = NULL)
```
In general, exponential smoothing is a class of forecasting procedures that rely on simple, recursive updating equations to compute forecasts. In particular, Simple ES is used for non-seasonal or deseasonalized time series that show no systematic trend. The updating equation is such that, given the N-th observation, the next prediction is a weighted average of the current observation and its forecast. Alpha is a smoothing parameter that controls the procedure's adaptiveness: the closer to 1 it is, the higher the weight is given to each observation in forecasting the next. Formally:
$$
\hat{Y}_{N+1} = \alpha Y_N +(1 - \alpha) \hat{Y}_{N}
$$

After deseasonalizing the time series, we have decided to perform the Holt-Winters forecasting procedure, instead of simple ES, in order to handle the very prominent trend component. The procedure thus produces estimates as a linear function of the estimation horizon:
$$
\textbf{Level update} \quad \quad
L_t = \alpha Y_t + (1 - \alpha) \left(L_{t-1} + T_{t-1} \right)
$$

$$
\textbf{Trend update} \quad \quad
T_t = \beta \left (L_t - L_{t-1} \right) + (1 - \beta) T_{t-1}
$$
$$
\textbf{h-step-ahead forecast at time t} \quad \quad
\hat{Y}_t (h) = (L_t + h T_t)
$$

We have let the model itself compute values for the smoothing parameters. Given a particular value of $\alpha$, $\beta$, the algorithm involves iteratively producing one-step-ahead forecasts throughout the series, and then computing the sum of squares of all the prediction errors. This can be repeated for different values of $\alpha$ according to a grid search or some other hyper-parameter tuning method, so that the value which minimizes the sum of squares is eventually found.

Our analysis leads us to use $\alpha = .44$, which means that each observation has a 44% weight on the next level estimate. If it were higher, then the line representing the one-step-ahead forecast in Figure$~$\ref{fig:hw} would be even more closely resembling the original data; if it were lower, then the blue line would be smoother.

### Forecasting the series

```{r, fig.height = 3, fig.width = 6, fig.show = "hold", out.width = "60%", fig.cap = "\\label{fig:hw_pred}15-year Holt–Winters forecast of deseasonalized series (2011–2025) with 80% and 95% confidence bands."}

actual <- window(deseasonalized, start = c(2011, 1), end = c(2025, 12))

# 15-year forecast from end of 2010
fit_holtwinters_forecast <- forecast(fit_holtwinters, h = 15 * 12)

fc_df <- data.frame(
  time = as.numeric(time(fit_holtwinters_forecast$mean)),
  mean = as.numeric(fit_holtwinters_forecast$mean),
  lo80 = as.numeric(fit_holtwinters_forecast$lower[, 1]),
  hi80 = as.numeric(fit_holtwinters_forecast$upper[, 1]),
  lo95 = as.numeric(fit_holtwinters_forecast$lower[, 2]),
  hi95 = as.numeric(fit_holtwinters_forecast$upper[, 2])
)

train_df <- data.frame(
  time  = as.numeric(time(train)),
  value = as.numeric(train)
)

actual_df <- data.frame(
  time = as.numeric(time(actual)),
  value = as.numeric(actual)
)

ggplot() +
  geom_line(data = train_df, aes(x = time, y = value, color = "Observed"), linewidth = 0.5) +
  # geom_line(data = df_hw(fit_holtwinters, train), aes(x = t, y = fitted, color = "One-step-ahead forecast")) +
  scale_color_manual(
    values = c(
      "Observed" = "black",
      "Forecast" = "red"
      )
    ) +
  scale_alpha_manual(
    values = c(
      "95% CI" = 0.2,
      "80% CI" = 0.3
    )
  ) +
  theme(legend.position = "inside", legend.position.inside = c(0.1, 0.8)) +
  labs(color = NULL) +
  geom_line(data = actual_df, aes(x = time, y = value), colour = "black", linewidth = 0.8) +
  geom_ribbon(data = fc_df, aes(x = time, ymin = lo95, ymax = hi95, alpha = "95% CI"), fill = "steelblue") +
  geom_ribbon(data = fc_df, aes(x = time, ymin = lo80, ymax = hi80, alpha = "80% CI"), fill = "steelblue") +
  geom_line(data = fc_df, aes(x = time, y = mean, colour = "Forecast"), linewidth = 0.8) +
  labs(
    title = NULL,
    x = NULL,
    y = "Temperature anomalies (°C)"
  ) +
  geom_hline(yintercept = 0)

```

```{r}
# MAE at horizons: 1, 5, 10, 15 years
forecasts <- fit_holtwinters_forecast$mean  # point forecasts from holt

horizons <- c(1, 5, 10, 15)
mae_results <- sapply(horizons, function(yr) {
  h <- yr * 12
  mean(abs(forecasts[1:h] - actual[1:h]))
})

mae_table <- data.frame(
  Horizon = horizons,
  MAE = mae_results
)

mape_results <- sapply(horizons, function(yr) {
  h <- yr * 12
  mean(abs(forecasts[1:h] - actual[1:h])) / mean(actual[1:h])
})

# kable(mae_table, digits = 7, align = c("r", "r"),
            #  caption = "MAE at increasing forecast horizons")

```

Figure$~$\ref{fig:hw_pred} includes the required plot.

As expected and reasonable, the mean absolute error increases with the horizon, h: it is equal to $0.10$ centigrade for a one-year horizon; $0.14$ for a 5-year horizon; $0.16$ for a 10-year horizon, and $0.18$ for a 15-year horizon.
