---
title: "Bocconi University - 20236 Time Series Analysis"
subtitle: "**Assignment 1 towards the final project | Group 27**"
output:
  pdf_document:
    latex_engine: xelatex
fontsize: 10pt
geometry: margin=2cm
---

```{r setup, include=FALSE}
rm(list = ls())
library(forecast)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

### Exploratory Analysis

```{r, fig.height=4, fig.width=6}
df <- read.csv("gistemp.csv", skip = 1, na.strings = "***")
#head(df)
monthly_matrix <- as.matrix(df[, 2:13])
monthly_vector <- as.numeric(t(monthly_matrix))

gistemp_ts <- ts(monthly_vector, start = c(1880, 1), frequency = 12)
plot(gistemp_ts, main = "Monthly Northern Hemisphere temperature anomalies (GISTEMP)", xlab = "Year", ylab = "Temperature anomalies  (°C)", xaxt = "n")
axis(1, at = seq(1880, 2020, by = 20))
abline(h = 0, lty = 2, col='red')
```

The time series shows a long-term upward trend in temperature anomalies over the sample period. Values are mostly negative before the mid-20th century and increase gradually over time. From approximately the 1980s onward, the rise becomes more pronounced, suggesting a possible acceleration in warming. Short-term fluctuations remain visible around this trend.

The marginal variances appear relatively stable, as the dispersion does not increase with the level of the series. There is no strong visual evidence of heteroskedasticity. The variance can therefore be considered approximately constant over time. This observation supports the use of an additive decomposition rather than a multiplicative one.

### Task 1

```{r task1_decompo, fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyr)
library(gridExtra)

#to ensure positive values when needed
shift <- 0
if (min(gistemp_ts, na.rm = TRUE) <= 0) {
  shift <- abs(min(gistemp_ts, na.rm = TRUE)) + 0.01
}
gistemp_pos <- gistemp_ts + shift
```


```{r task1_decompo, fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
dec_add  <- decompose(gistemp_ts,  type = "additive")
dec_mult <- decompose(gistemp_pos, type = "multiplicative")
stl_fit  <- stl(gistemp_ts, s.window = "periodic")

#conversion of the decompose() object for a cleaner output
df_decomp <- function(dec_obj) {
  data.frame(
    t = as.numeric(time(dec_obj$x)),
    month = as.integer(cycle(dec_obj$x)), #monthly index to aggregate seasonal effects for better visualization
    observed = as.numeric(dec_obj$x),
    trend = as.numeric(dec_obj$trend),
    seasonal = as.numeric(dec_obj$seasonal),
    residual = as.numeric(dec_obj$random)
  )
}

df_stl <- function(stl_obj, y) {
  data.frame(
    t = as.numeric(time(stl_obj$time.series[, 1])),
    month = as.integer(cycle(y)), #monthly index as in df_decomp
    observed = as.numeric(y),
    trend = as.numeric(stl_obj$time.series[, "trend"]),
    seasonal = as.numeric(stl_obj$time.series[, "seasonal"]),
    residual = as.numeric(stl_obj$time.series[, "remainder"])
  )
}

#formatting as a panel using subplots
panel <- function(df, main_title) {
  base_theme <- theme_minimal(base_size = 10) +
    theme(
      plot.title = element_text(face = "bold", size = 10),
      panel.grid.minor = element_blank()
    )

  p_observed <- ggplot(df, aes(t, observed)) + geom_line(size = 0.3) +
    labs(title = "Observed", x = NULL, y = NULL) + base_theme

  p_trend <- ggplot(df, aes(t, trend)) + geom_line(size = 0.3) +
    labs(title = "Trend", x = NULL, y = NULL) + base_theme

  #aggregation for the seasonality : we create average seasonal profile over the 12 months
  df_seasonal <- aggregate(seasonal ~ month, data = df, FUN = mean, na.rm = TRUE)
  p_seasonal <- ggplot(df_seasonal, aes(month, seasonal)) +
    geom_line(size = 0.4) + geom_point(size = 0.8) +
    scale_x_continuous(
      breaks = 1:12,
      labels = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
    ) + labs(title = "Seasonal", x = NULL, y = NULL) + base_theme

  p_residual <- ggplot(df, aes(t, residual)) + geom_line(size = 0.3) +
    labs(title = "Residual", x = NULL, y = NULL) + base_theme

  # Arrange in a 2x2 grid and RETURN it (important for knitting)
  gridExtra::grid.arrange(
    p_observed, p_trend,
    p_seasonal, p_residual,
    ncol = 2,
    top = grid::textGrob(main_title, gp = grid::gpar(fontface = "bold", fontsize = 14))
  )
}

#using grid to display panels
grid::grid.draw(panel(df_decomp(dec_add),  "Additive decomposition (GISTEMP)"))
grid::grid.draw(panel(df_decomp(dec_mult), "Multiplicative decomposition (shifted GISTEMP)"))
grid::grid.draw(panel(df_stl(stl_fit, gistemp_ts), "STL decomposition (periodic seasonality)"))
```


### Task 2

```{r}
seasonal_component <- stl_fit$time.series[, "seasonal"]
deseasonalized <- gistemp_ts - seasonal_component
train <- window(deseasonalized, start = c(1950, 1), end = c(2010, 12))

par(mar = c(4,4,2,1))  # cleaner margins

plot(train,
     main = "Deseasonalized Northern Hemisphere temperature anomalies (1950–2010)",
     xlab = "Year",
     ylab = "Temperature anomaly (°C)",
     col = "black",
     lwd = 1,
     cex.main = 1,
     cex.lab = 0.9,
     cex.axis = 0.8)

abline(h = 0, lty = 2, col = "red")

```


```{r}

library(forecast)

## --- Fit Holt's linear method (beta = TRUE, gamma = FALSE) ---
fit_holt <- holt(train, h = 1)

alpha_holt <- fit_holt$model$par["alpha"]
beta_holt  <- fit_holt$model$par["beta"]

alpha_holt  # 0.288203
beta_holt   # 0.0003834748

## --- One-step-ahead fitted values vs observed ---
par(mar = c(4, 4, 2, 1))

plot(train,
     main     = "Holt one-step-ahead forecasts vs observed (1950–2010)",
     xlab     = "Year",
     ylab     = "Temperature anomaly (°C)",
     col      = "black",
     lwd      = 1,
     cex.main = 1,
     cex.lab  = 0.9,
     cex.axis = 0.8)

lines(fitted(fit_holt), col = "red", lwd = 1.2)

legend("topleft",
       legend = c("Observed", "One-step-ahead forecast"),
       col    = c("black", "red"),
       lwd    = c(1, 1.2),
       cex    = 0.8,
       bty    = "n")



```

```{r}
## --- Holt-Winters with trend (beta = TRUE, gamma = FALSE) ---
fit_holt2 <- HoltWinters(train, beta = TRUE, gamma = FALSE)

# Extract parameters correctly
alpha_holt2 <- fit_holt2$alpha   # numeric estimated alpha
beta_holt2  <- fit_holt2$beta    # numeric estimated beta — this is the fix!

alpha_holt2
beta_holt2

## --- One-step-ahead fitted values vs observed ---
par(mar = c(4, 4, 2, 1))

plot(train,
     main    = "Holt-Winters one-step-ahead forecasts vs observed (1950–2010)",
     xlab    = "Year",
     ylab    = "Temperature anomaly (°C)",
     col     = "black",
     lwd     = 1,
     cex.main = 1,
     cex.lab  = 0.9,
     cex.axis = 0.8)

lines(fitted(fit_holt2)[, "xhat"], col = "red", lwd = 1.2)

legend("topleft",
       legend = c("Observed", "One-step-ahead forecast"),
       col    = c("black", "red"),
       lwd    = c(1, 1.2),
       cex    = 0.8,
       bty    = "n")

```

```{r}
# Forecasting the deseasonalized series (15 years ahead = 180 months)
library(forecast)

h_years <- 15
h_months <- h_years * 12

# --- Refit/compute forecasts with holt() (forecast package) on the deseasonalized train ---
# (train is already the deseasonalized ts for Jan1950-Dec2010 from your previous code)
holt_fc <- holt(train, h = h_months, level = c(80, 95), damped = FALSE)

# -- Forecast from your HoltWinters fit (already named fit_holt2) --
# fit_holt2 was fitted to 'train' earlier via HoltWinters(train, beta=TRUE, gamma=FALSE)
hw_fc <- forecast(fit_holt2, h = h_months, level = c(80, 95))

# --- Plot deseasonalized observed (train) and forecasts (both methods) ---
par(mfrow = c(1,2), mar = c(4,4,2,1))

# Panel A: holt() forecast (deseasonalized)
plot(window(deseasonalized, start = c(1950,1), end = c(2010,12)),
     main = "Holt forecasts on deseasonalized series (1950–2010 train)",
     xlab = "Year", ylab = "Deseasonalized anomaly (°C)",
     col = "black", lwd = 1)
# add holt point forecasts and intervals
lines(holt_fc$mean, col = "red", lwd = 1.2)
# 80% interval
lines(ts(holt_fc$lower[ ,1], start = time(holt_fc$mean)[1], frequency = 12), col = "red", lty = 2)
lines(ts(holt_fc$upper[ ,1], start = time(holt_fc$mean)[1], frequency = 12), col = "red", lty = 2)
# 95% interval
lines(ts(holt_fc$lower[ ,2], start = time(holt_fc$mean)[1], frequency = 12), col = "red", lty = 3)
lines(ts(holt_fc$upper[ ,2], start = time(holt_fc$mean)[1], frequency = 12), col = "red", lty = 3)

legend("topleft", legend = c("Observed (train)", "Holt mean", "80% PI", "95% PI"),
       col = c("black", "red", "red", "red"), lwd = c(1,1.2,1,1), lty = c(1,1,2,3),
       bty = "n", cex = 0.8)

# Panel B: Holt-Winters (HoltWinters fitted to train) forecast
plot(window(deseasonalized, start = c(1950,1), end = c(2010,12)),
     main = "Holt-Winters (via forecast()) on deseasonalized series",
     xlab = "Year", ylab = "Deseasonalized anomaly (°C)",
     col = "black", lwd = 1)
lines(hw_fc$mean, col = "blue", lwd = 1.2)
lines(ts(hw_fc$lower[ ,1], start = time(hw_fc$mean)[1], frequency = 12), col = "blue", lty = 2)
lines(ts(hw_fc$upper[ ,1], start = time(hw_fc$mean)[1], frequency = 12), col = "blue", lty = 2)
lines(ts(hw_fc$lower[ ,2], start = time(hw_fc$mean)[1], frequency = 12), col = "blue", lty = 3)
lines(ts(hw_fc$upper[ ,2], start = time(hw_fc$mean)[1], frequency = 12), col = "blue", lty = 3)

legend("topleft", legend = c("Observed (train)", "HW mean", "80% PI", "95% PI"),
       col = c("black", "blue", "blue", "blue"), lwd = c(1,1.2,1,1), lty = c(1,1,2,3),
       bty = "n", cex = 0.8)

# reset plotting
par(mfrow = c(1,1))

# --- MAE evaluation at horizons if observed out-of-sample exists (deseasonalized) ---
# observed out-of-sample deseasonalized values (2011-...): may be shorter than full 15 years depending on your CSV
obs_oos <- window(deseasonalized, start = c(2011,1), end = c(2010 + h_years, 12))

horizons <- c(1, 5, 10, 15)
mae_table <- data.frame(horizon_years = horizons, MAE_holt = NA_real_, MAE_hw = NA_real_)

for (i in seq_along(horizons)) {
  yrs <- horizons[i]
  m <- yrs * 12
  # ensure we have at least m observed months after 2010
  if (length(obs_oos) >= m) {
    obs_slice <- as.numeric(window(deseasonalized, start = c(2011,1), end = c(2010 + yrs, 12)))
    holt_slice <- as.numeric(holt_fc$mean[1:m])
    hw_slice   <- as.numeric(hw_fc$mean[1:m])
    mae_table$MAE_holt[i] <- mean(abs(obs_slice - holt_slice))
    mae_table$MAE_hw[i]   <- mean(abs(obs_slice - hw_slice))
  } else {
    mae_table$MAE_holt[i] <- NA
    mae_table$MAE_hw[i]   <- NA
  }
}

# write MAE table for the report (and print to console for quick check)
write.csv(mae_table, "mae_task2_holt_vs_hw.csv", row.names = FALSE)
mae_table
```

```{r}
## --- 15-year forecast from end of 2010 ---
fit_holt_forecast <- holt(train, h = 15 * 12)

autoplot(fit_holt_forecast) +
  autolayer(train, series = "Observed") +
  labs(
    title  = "15-year Holt forecast of deseasonalized temperature anomalies (2011–2025)",
    x      = "Year",
    y      = "Temperature anomaly (°C)"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.title = element_blank())
```


```{r}
library(forecast)
library(ggplot2)

## --- 15-year forecast from end of 2010 ---
fit_holt_forecast <- forecast(holt(train, h = 15 * 12))

fc_df <- data.frame(
  time = as.numeric(time(fit_holt_forecast$mean)),
  mean = as.numeric(fit_holt_forecast$mean),
  lo80 = as.numeric(fit_holt_forecast$lower[, 1]),
  hi80 = as.numeric(fit_holt_forecast$upper[, 1]),
  lo95 = as.numeric(fit_holt_forecast$lower[, 2]),
  hi95 = as.numeric(fit_holt_forecast$upper[, 2])
)

train_df <- data.frame(
  time  = as.numeric(time(train)),
  value = as.numeric(train)
)

ggplot() +
  geom_ribbon(data = fc_df, aes(x = time, ymin = lo95, ymax = hi95), fill = "steelblue", alpha = 0.2) +
  geom_ribbon(data = fc_df, aes(x = time, ymin = lo80, ymax = hi80), fill = "steelblue", alpha = 0.3) +
  geom_line(data = train_df, aes(x = time, y = value), colour = "black", linewidth = 0.5) +
  geom_line(data = fc_df, aes(x = time, y = mean), colour = "steelblue", linewidth = 0.8) +
  labs(
    title = "15-year Holt forecast of deseasonalized temperature anomalies (2011–2025)",
    x     = "Year",
    y     = "Temperature anomaly (°C)"
  ) +
  theme_minimal(base_size = 10)
```
```{r}
## --- MAE at horizons: 1, 5, 10, 15 years ---
actual <- window(deseasonalized, start = c(2011, 1), end = c(2025, 12))
forecasts <- fit_holt_forecast$mean  # point forecasts from holt

horizons <- c(1, 5, 10, 15)
mae_results <- sapply(horizons, function(yr) {
  h <- yr * 12
  mean(abs(forecasts[1:h] - actual[1:h]))
})

mae_table <- data.frame(
  horizon_years = horizons,
  MAE_holt      = mae_results
)

knitr::kable(mae_table, digits = 7, align = c("r", "r"),
             caption = "MAE at increasing forecast horizons")

```
