---
title: "Bocconi University - 20236 Time Series Analysis"
subtitle: "**Assignment 1 towards the final project | Group 27**"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: yes
    extra_dependencies: ["float"]
fontsize: 10pt
geometry: margin=2cm
---

```{r setup, include=FALSE}
rm(list = ls())
library(forecast)
library(tidyverse)
library(gridExtra)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center", fig.pos = "H", out.extra = "")

plot_ts_gg <- function(
    x,
    title = NULL,
    ylab = NULL,
    aes = NULL,
    facet_ncol = 1,
    wrap_width = 45
) {
  if (!inherits(x, "ts")) {
    stop("Input must be ts.")
  }

  df <- tibble(
    Time = as.numeric(time(x)),
    Value = as.numeric(x)
  )

  p <- ggplot(df, aes(Time, Value, !!!aes)) +
    geom_line() +
    labs(title = title, x = NULL, y = ylab)+
    theme_minimal(base_size = 10)

  p
}
```

### Exploratory Analysis

```{r, fig.height=3, fig.width=6, fig.cap = "\\label{fig:main}Monthly Northern Hemisphere temperature anomalies (GISTEMP)."}
df <- read.csv("gistemp.csv", skip = 1, na.strings = "***")
#head(df)
monthly_matrix <- as.matrix(df[, 2:13])
monthly_vector <- as.numeric(t(monthly_matrix))

gistemp_ts <- ts(monthly_vector, start = c(1880, 1), frequency = 12)

plot_ts_gg(
  gistemp_ts,
  ylab = "Temperature anomalies (°C)"
) +
  geom_hline(yintercept = 0)
```

The time series (Figure$~$\ref{fig:main}) shows a long-term upward trend in temperature anomalies over the sample period. Values are mostly negative before the mid-20th century and increase gradually over time. From approximately the 1980s onward, the rise becomes more pronounced, suggesting a possible acceleration in warming. Short-term fluctuations remain visible around this trend.

The marginal variances appear relatively stable, as the dispersion does not increase with the level of the series. There is no strong visual evidence of heteroskedasticity. The variance can therefore be considered approximately constant over time. This observation supports the use of an additive decomposition rather than a multiplicative one.

### Task 1

```{r task1_decompo_prep, fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
#to ensure positive values when needed
shift <- 0
if (min(gistemp_ts, na.rm = TRUE) <= 0) {
  shift <- abs(min(gistemp_ts, na.rm = TRUE)) + 0.01
}
gistemp_pos <- gistemp_ts + shift
```


```{r task1_decompo, fig.width=10, fig.height=4, out.width = "100%", fig.cap = "\\label{fig:decomp}Comparison of time-series decompositions of Northern Hemisphere temperature anomalies.", fig.show = "hold"}
dec_add  <- decompose(gistemp_ts,  type = "additive")
dec_mult <- decompose(gistemp_pos, type = "multiplicative")
stl_fit  <- stl(gistemp_ts, s.window = "periodic")

# Assuming same seasonality for all decompositions
const_seasonality <- frequency(gistemp_ts)

#conversion of the decompose() object for a cleaner output
df_decomp <- function(dec_obj) {
  data.frame(
    t = as.numeric(time(dec_obj$x)),
    observed = as.numeric(dec_obj$x),
    trend = as.numeric(dec_obj$trend),
    seasonal = as.numeric(dec_obj$seasonal),
    residual = as.numeric(dec_obj$random)
  )
}

df_stl <- function(stl_obj, y) {
  t <- as.numeric(time(stl_obj$time.series[, 1])) #extracting time index
  data.frame(
    t = t,
    observed = as.numeric(y),
    trend = as.numeric(stl_obj$time.series[, "trend"]),
    seasonal = as.numeric(stl_obj$time.series[, "seasonal"]),
    residual = as.numeric(stl_obj$time.series[, "remainder"])
  )
}

df_hw <- function(hw_obj, y, pad = 2) {
  t <- as.numeric(time(y)) #extracting time index
  tibble(
    t = t,
    observed = as.numeric(y),
    fitted = c(rep(NA, 2), as.numeric(fitted(hw_obj)[, "xhat"])),
    trend = c(rep(NA, 2), as.numeric(fitted(hw_obj)[, "level"]) + as.numeric(fitted(hw_obj)[, "trend"])),
    residual = observed - fitted
  )
}

#formatting as a panel using subplots
panel <- function(df, main_title) {
  base_theme <- theme_minimal(base_size = 10) +
    theme(
      plot.title = element_text(face = "bold", size = 10),
      panel.grid.minor = element_blank()
    )

  p_observed <- ggplot(df, aes(t, observed)) + geom_line(size = 0.3) +
    labs(title = "Observed", x = NULL, y = NULL) + base_theme

  p_trend <- ggplot(df, aes(t, trend)) + geom_line(size = 0.3) +
    labs(title = "Trend", x = NULL, y = NULL) + base_theme

  seasonal_seq <- seq(from = 1, to = const_seasonality)

  p_seasonal <- df %>%
    slice_head(n = const_seasonality) %>%
    mutate(t = seasonal_seq) %>%
    ggplot(aes(t, seasonal)) + geom_line(size = 0.3) +
    scale_x_continuous(breaks = seasonal_seq) +
    labs(title = "Seasonal", x = NULL, y = NULL) + base_theme


  p_residual <- ggplot(df, aes(t, residual)) + geom_line(size = 0.3) +
    labs(title = "Residual", x = NULL, y = NULL) + base_theme

  # Arrange in a 2x2 grid and RETURN it (important for knitting)
  gridExtra::grid.arrange(
    p_observed, p_trend,
    p_seasonal, p_residual,
    ncol = 2,
    top = grid::textGrob(main_title, gp = grid::gpar(fontface = "bold", fontsize = 14))
  )
}

#using grid to display panels
grid::grid.draw(panel(df_decomp(dec_add),  "Additive decomposition (GISTEMP)"))
grid::grid.draw(panel(df_decomp(dec_mult), "Multiplicative decomposition (shifted GISTEMP)"))
grid::grid.draw(panel(df_stl(stl_fit, gistemp_ts), "STL decomposition (periodic seasonality)"))
```


### Task 2

```{r, fig.width = 6, fig.height = 3, fig.show = "hold", fig.cap = "\\label{fig:hw}Deseasonalized Northern Hemisphere temperature anomalies (1950–2010) and Holt–Winters one-step ahead forceasts."}
seasonal_component <- stl_fit$time.series[, "seasonal"]
deseasonalized <- gistemp_ts - seasonal_component
train <- window(deseasonalized, start = c(1950, 1), end = c(2010, 12))

# plot_ts_gg(train,
#   ylab = "Temperature anomaly (°C)",
# ) +
#   geom_hline(yintercept = 0)

## Holt-Winters with trend (gamma = FALSE)
fit_holtwinters <- HoltWinters(train, gamma = FALSE, optim.start = c(alpha = 0.8, beta = 0.9))

# Extract parameters correctly
alpha_holtwinters <- fit_holtwinters$alpha
beta_holtwinters  <- fit_holtwinters$beta

alpha_holtwinters
beta_holtwinters

plot_ts_gg(train,
  ylab = "Temperature anomaly (°C)",
  aes = list(color = "Observed"),
) +
  geom_hline(yintercept = 0) +
  geom_line(data = df_hw(fit_holtwinters, train), aes(x = t, y = fitted, color = "One-step-ahead forecast")) +
  scale_color_manual(values = c("black", "red")) +
  theme(legend.position = "inside", legend.position.inside = c(0.2, 0.9)) +
  labs(color = NULL)
```





```{r, fig.width = 6, fig.height = 3, fig.cap = "\\label{fig:hw_pred}15-year Holt–Winters forecast of deseasonalized temperature anomalies (2011–2025) with 80% and 95% confidence bands."}
## --- 15-year forecast from end of 2010 ---
fit_holtwinters_forecast <- forecast(fit_holtwinters, h = 15 * 12)

fc_df <- data.frame(
  time = as.numeric(time(fit_holtwinters_forecast$mean)),
  mean = as.numeric(fit_holtwinters_forecast$mean),
  lo80 = as.numeric(fit_holtwinters_forecast$lower[, 1]),
  hi80 = as.numeric(fit_holtwinters_forecast$upper[, 1]),
  lo95 = as.numeric(fit_holtwinters_forecast$lower[, 2]),
  hi95 = as.numeric(fit_holtwinters_forecast$upper[, 2])
)

train_df <- data.frame(
  time  = as.numeric(time(train)),
  value = as.numeric(train)
)

ggplot() +
  geom_ribbon(data = fc_df, aes(x = time, ymin = lo95, ymax = hi95), fill = "steelblue", alpha = 0.2) +
  geom_ribbon(data = fc_df, aes(x = time, ymin = lo80, ymax = hi80), fill = "steelblue", alpha = 0.3) +
  geom_line(data = train_df, aes(x = time, y = value), colour = "black", linewidth = 0.5) +
  geom_line(data = fc_df, aes(x = time, y = mean), colour = "steelblue", linewidth = 0.8) +
  labs(
    x     = "Year",
    y     = "Temperature anomaly (°C)"
  ) +
  theme_minimal(base_size = 10) +
  geom_hline(yintercept = 0)
```
```{r}
## --- MAE at horizons: 1, 5, 10, 15 years ---
actual <- window(deseasonalized, start = c(2011, 1), end = c(2025, 12))
forecasts <- fit_holtwinters_forecast$mean  # point forecasts from holt

horizons <- c(1, 5, 10, 15)
mae_results <- sapply(horizons, function(yr) {
  h <- yr * 12
  mean(abs(forecasts[1:h] - actual[1:h]))
})

mae_table <- data.frame(
  horizon_years = horizons,
  MAE_holtwinters = mae_results
)

knitr::kable(mae_table, digits = 7, align = c("r", "r"),
             caption = "MAE at increasing forecast horizons")

```
