---
title: "Bocconi University - 20236 Time Series Analysis"
subtitle: "**Assignment 1 towards the final project | Group 27**"
output:
  pdf_document:
    latex_engine: xelatex
fontsize: 10pt
geometry: margin=2cm
---

```{r setup, include=FALSE}
rm(list = ls())
library(forecast)
library(ggplot2)
library(tidyr)
library(gridExtra)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

### Exploratory Analysis

```{r, fig.height=4, fig.width=6}
df <- read.csv("gistemp.csv", skip = 1, na.strings = "***")
#head(df)
monthly_matrix <- as.matrix(df[, 2:13])
monthly_vector <- as.numeric(t(monthly_matrix))

gistemp_ts <- ts(monthly_vector, start = c(1880, 1), frequency = 12)
plot(gistemp_ts, main = "Monthly Northern Hemisphere temperature anomalies (GISTEMP)", xlab = "Year", ylab = "Temperature anomalies  (°C)", xaxt = "n")
axis(1, at = seq(1880, 2020, by = 20))
abline(h = 0, lty = 2, col='red')
```

The time series shows a long-term upward trend in temperature anomalies over the sample period. Values are mostly negative before the mid-20th century and increase gradually over time. From approximately the 1980s onward, the rise becomes more pronounced, suggesting a possible acceleration in warming. Short-term fluctuations remain visible around this trend.

The marginal variances appear relatively stable, as the dispersion does not increase with the level of the series. There is no strong visual evidence of heteroskedasticity. The variance can therefore be considered approximately constant over time. This observation supports the use of an additive decomposition rather than a multiplicative one.

### Task 1

```{r task1_decompo, fig.width=10, fig.height=8, message=FALSE, warning=FALSE}


#to ensure positive values when needed
shift <- 0
if (min(gistemp_ts, na.rm = TRUE) <= 0) {
  shift <- abs(min(gistemp_ts, na.rm = TRUE)) + 0.01
}
gistemp_pos <- gistemp_ts + shift
```


```{r task1_decompo, fig.width=10, fig.height=8, message=FALSE, warning=FALSE}
dec_add  <- decompose(gistemp_ts,  type = "additive")
dec_mult <- decompose(gistemp_pos, type = "multiplicative")
stl_fit  <- stl(gistemp_ts, s.window = "periodic")

#conversion of the decompose() object for a cleaner output
df_decomp <- function(dec_obj) {
  data.frame(
    t = as.numeric(time(dec_obj$x)),
    observed = as.numeric(dec_obj$x),
    trend = as.numeric(dec_obj$trend),
    seasonal = as.numeric(dec_obj$seasonal),
    residual = as.numeric(dec_obj$random)
  )
}

df_stl <- function(stl_obj, y) {
  t <- as.numeric(time(stl_obj$time.series[, 1])) #extracting time index 
  data.frame(
    t = t,
    observed = as.numeric(y),
    trend = as.numeric(stl_obj$time.series[, "trend"]),
    seasonal = as.numeric(stl_obj$time.series[, "seasonal"]),
    residual = as.numeric(stl_obj$time.series[, "remainder"])
  )
}

#formatting as a panel using subplots
panel <- function(df, main_title) {
  base_theme <- theme_minimal(base_size = 10) +
    theme(
      plot.title = element_text(face = "bold", size = 10),
      panel.grid.minor = element_blank()
    )

  p_observed <- ggplot(df, aes(t, observed)) + geom_line(size = 0.3) +
    labs(title = "Observed", x = NULL, y = NULL) + base_theme

  p_trend <- ggplot(df, aes(t, trend)) + geom_line(size = 0.3) +
    labs(title = "Trend", x = NULL, y = NULL) + base_theme

  p_seasonal <- ggplot(df, aes(t, seasonal)) + geom_line(size = 0.3) +
    labs(title = "Seasonal", x = NULL, y = NULL) + base_theme

  p_residual <- ggplot(df, aes(t, residual)) + geom_line(size = 0.3) +
    labs(title = "Residual", x = NULL, y = NULL) + base_theme

  # Arrange in a 2x2 grid and RETURN it (important for knitting)
  gridExtra::grid.arrange(
    p_observed, p_trend,
    p_seasonal, p_residual,
    ncol = 2,
    top = grid::textGrob(main_title, gp = grid::gpar(fontface = "bold", fontsize = 14))
  )
}

#using grid to display panels
grid::grid.draw(panel(df_decomp(dec_add),  "Additive decomposition (GISTEMP)"))
grid::grid.draw(panel(df_decomp(dec_mult), "Multiplicative decomposition (shifted GISTEMP)"))
grid::grid.draw(panel(df_stl(stl_fit, gistemp_ts), "STL decomposition (periodic seasonality)"))
```


### Task 2

```{r}
seasonal_component <- stl_fit$time.series[, "seasonal"]
deseasonalized <- gistemp_ts - seasonal_component
train <- window(deseasonalized, start = c(1950, 1), end = c(2010, 12))

par(mar = c(4,4,2,1))  # cleaner margins

plot(train,
     main = "Deseasonalized Northern Hemisphere temperature anomalies (1950–2010)",
     xlab = "Year",
     ylab = "Temperature anomaly (°C)",
     col = "black",
     lwd = 1,
     cex.main = 1,
     cex.lab = 0.9,
     cex.axis = 0.8)

abline(h = 0, lty = 2, col = "red")

```



```{r}
## Holt-Winters with trend (gamma = FALSE)
fit_holtwinters <- HoltWinters(train, gamma = FALSE)

# Extract parameters correctly
alpha_holtwinters <- fit_holtwinters$alpha   # numeric estimated alpha
beta_holtwinters  <- fit_holtwinters$beta    # numeric estimated beta — this is the fix!

alpha_holtwinters
beta_holtwinters

# One-step-ahead fitted values vs observed 
par(mar = c(4, 4, 2, 1))

plot(train,
     main    = "Holt-Winters one-step-ahead forecasts vs observed (1950–2010)",
     xlab    = "Year",
     ylab    = "Temperature anomaly (°C)",
     col     = "black",
     lwd     = 1,
     cex.main = 1,
     cex.lab  = 0.9,
     cex.axis = 0.8)

lines(fitted(fit_holtwinters)[, "xhat"], col = "red", lwd = 1.2)

legend("topleft",
       legend = c("Observed", "One-step-ahead forecast"),
       col    = c("black", "red"),
       lwd    = c(1, 1.2),
       cex    = 0.8,
       bty    = "n")

```





```{r}
## --- 15-year forecast from end of 2010 ---
fit_holtwinters_forecast <- forecast(HoltWinters(train, gamma = FALSE), h = 15 * 12)

fc_df <- data.frame(
  time = as.numeric(time(fit_holtwinters_forecast$mean)),
  mean = as.numeric(fit_holtwinters_forecast$mean),
  lo80 = as.numeric(fit_holtwinters_forecast$lower[, 1]),
  hi80 = as.numeric(fit_holtwinters_forecast$upper[, 1]),
  lo95 = as.numeric(fit_holtwinters_forecast$lower[, 2]),
  hi95 = as.numeric(fit_holtwinters_forecast$upper[, 2])
)

train_df <- data.frame(
  time  = as.numeric(time(train)),
  value = as.numeric(train)
)

ggplot() +
  geom_ribbon(data = fc_df, aes(x = time, ymin = lo95, ymax = hi95), fill = "steelblue", alpha = 0.2) +
  geom_ribbon(data = fc_df, aes(x = time, ymin = lo80, ymax = hi80), fill = "steelblue", alpha = 0.3) +
  geom_line(data = train_df, aes(x = time, y = value), colour = "black", linewidth = 0.5) +
  geom_line(data = fc_df, aes(x = time, y = mean), colour = "steelblue", linewidth = 0.8) +
  labs(
    title = "15-year Holt forecast of deseasonalized temperature anomalies (2011–2025)",
    x     = "Year",
    y     = "Temperature anomaly (°C)"
  ) +
  theme_minimal(base_size = 10)
```
```{r}
## --- MAE at horizons: 1, 5, 10, 15 years ---
actual <- window(deseasonalized, start = c(2011, 1), end = c(2025, 12))
forecasts <- fit_holtwinters_forecast$mean  # point forecasts from holt

horizons <- c(1, 5, 10, 15)
mae_results <- sapply(horizons, function(yr) {
  h <- yr * 12
  mean(abs(forecasts[1:h] - actual[1:h]))
})

mae_table <- data.frame(
  horizon_years = horizons,
  MAE_holtwinters      = mae_results
)

knitr::kable(mae_table, digits = 7, align = c("r", "r"),
             caption = "MAE at increasing forecast horizons")

```
