---
title: "Bocconi University - 20236 Time Series Analysis"
subtitle: "**Assignment 1 towards the final project | Group 27**"
output:
  pdf_document:
    latex_engine: xelatex
fontsize: 10pt
geometry: margin=2cm
header-includes:
  - \usepackage{float}
  - \setlength{\textfloatsep}{6pt}
  - \setlength{\intextsep}{6pt}
---

```{r setup, include=FALSE}
rm(list = ls())
library(forecast)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

### Exploratory Analysis

```{r, fig.height=4, fig.width=6}
df <- read.csv("gistemp.csv", skip = 1, na.strings = "***")
#head(df)
monthly_matrix <- as.matrix(df[, 2:13])
monthly_vector <- as.numeric(t(monthly_matrix))

gistemp_ts <- ts(monthly_vector, start = c(1880, 1), frequency = 12)
plot(gistemp_ts, main = "Monthly Northern Hemisphere temperature anomalies (GISTEMP)", xlab = "Year", ylab = "Temperature anomalies  (°C)", xaxt = "n")
axis(1, at = seq(1880, 2020, by = 20))
abline(h = 0, lty = 2, col='red')
```

The time series shows a long-term upward trend in temperature anomalies over the sample period. Values are mostly negative before the mid-20th century and increase gradually over time. From approximately the 1980s onward, the rise becomes more pronounced, suggesting a possible acceleration in warming. Short-term fluctuations remain visible around this trend.

The marginal variances appear relatively stable, as the dispersion does not increase with the level of the series. There is no strong visual evidence of heteroskedasticity. The variance can therefore be considered approximately constant over time. This observation supports the use of an additive decomposition rather than a multiplicative one.

### Task 1

```{r task1_decompo, message=FALSE, warning=FALSE}
library(ggplot2)
library(gridExtra)

#to ensure positive values when needed
shift <- 0
if (min(gistemp_ts, na.rm = TRUE) <= 0) {
  shift <- abs(min(gistemp_ts, na.rm = TRUE)) + 0.01
}
gistemp_pos <- gistemp_ts + shift

dec_add  <- decompose(gistemp_ts,  type = "additive")
dec_mult <- decompose(gistemp_pos, type = "multiplicative")
stl_fit  <- stl(gistemp_ts, s.window = "periodic")

#conversion of the decompose() object for a cleaner output
df_decomp <- function(dec_obj) {
  data.frame(
    t = as.numeric(time(dec_obj$x)),
    month = as.integer(cycle(dec_obj$x)), #monthly index to aggregate seasonal effects for better visualization
    observed = as.numeric(dec_obj$x),
    trend = as.numeric(dec_obj$trend),
    seasonal = as.numeric(dec_obj$seasonal),
    residual = as.numeric(dec_obj$random)
  )
}

df_stl <- function(stl_obj, y) {
  data.frame(
    t = as.numeric(time(stl_obj$time.series[, 1])),
    month = as.integer(cycle(y)), #monthly index as in df_decomp
    observed = as.numeric(y),
    trend = as.numeric(stl_obj$time.series[, "trend"]),
    seasonal = as.numeric(stl_obj$time.series[, "seasonal"]),
    residual = as.numeric(stl_obj$time.series[, "remainder"])
  )
}

df_add  <- df_decomp(dec_add)
df_mult <- df_decomp(dec_mult)
df_s    <- df_stl(stl_fit, gistemp_ts)
```

```{r task1_trend, fig.width=10, fig.height=4, message=FALSE, warning=FALSE, fig.pos="H"}

op <- par(mfrow = c(1, 2))
plot(df_add$t, df_add$trend, type = "l", main = "Trend (Additive)", xlab = "Year", ylab = "")
plot(df_mult$t, df_mult$trend, type = "l", main = "Trend (Multiplicative)", xlab = "Year", ylab = "")
par(op)
```
The trend component clearly reveals a long-term warming pattern in the GISTEMP data. After relatively moderate fluctuations in the late 19th and early 20th centuries, temperatures begin to increase gradually. A temporary stabilization is visible around the mid-20th century, with a relatively flatter slope between 1940 and 1970 suggesting a slowdown in warming. However, from the late 1970s onward, the trend exhibits a much steeper slope, indicating a strong and persistent acceleration in warming.
The additive and multiplicative decompositions produce nearly identical trend components, indicating that the estimated long-term dynamics are robust to the choice of model.

```{r task1_seasonality, fig.width=10, fig.height=3.2, message=FALSE, warning=FALSE}
library(ggplot2)
library(gridExtra)

#theme to prevent overlap
base_theme <- theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold", size = 9),
        plot.margin = margin(2, 6, 2, 2),
        panel.grid.minor = element_blank())

ggplot(df_add, aes(t, seasonal)) +
  geom_line(size = 0.2) +
  labs(
    title = "Seasonal component over the full sample (dense monthly pattern)",
    x = "Year", y = NULL
  ) +
  base_theme
```
Because the series is monthly and spans more than a century, the seasonal component plotted over the full sample is visually dense and hard to interpret. For readability, we therefore zoom in on a recent subperiod (from 2000 onward), which preserves the seasonal structure while avoiding overplotting.

```{r task1_seasonality_zoom, fig.width=13, fig.height=6, message=FALSE, warning=FALSE}
library(ggplot2)
library(gridExtra)

df_add_z  <- subset(df_add,  t >= 2000) #from year 2000
df_mult_z <- subset(df_mult, t >= 2000)
df_s_z    <- subset(df_s,    t >= 2000)

mk_seas_time <- function(df, ttl, ref_line) {
  ggplot(df, aes(t, seasonal)) +
    geom_line(size = 0.25) +
    geom_hline(yintercept = ref_line, linetype = "dashed", size = 0.25, col='red') +
    labs(title = ttl, x = "Year", y = NULL) +
    base_theme +
    theme(plot.title = element_text(size = 9))
}

p_add  <- mk_seas_time(df_add_z, "Additive seasonality, since 2000",0)

p_stl  <- mk_seas_time(df_s_z, "STL seasonality (LOESS), since 2000",0)

p_mult <- mk_seas_time(df_mult_z, "Multiplicative seasonality, since 2000",1)

gridExtra::grid.arrange(p_add, p_stl, p_mult, ncol = 3)

```
The zoomed figures confirm that the seasonal pattern remains highly stable over time. We don not observe significant change in amplitude over the recent period, suggesting that seasonality is approximately constant. As, in the multiplicative decompositon, the seasonal component is expressed as a multiplicative factor, the plot is centered around 1, whereas in the additive (and STL) decomposition it is centered around 0.

To better understand the intra-annual structure of the seasonal component, we now examine the average seasonal profile by month. Aggregating across years provides a more interpretable representation of the annual cycle of the series.

```{r task1_seasonality_aggregate, fig.width=13, fig.height=6, message=FALSE, warning=FALSE}
library(ggplot2)
library(gridExtra)

#We create seasonal profile (12months average) for readability
agg_add  <- aggregate(seasonal ~ month, data = df_add,  FUN = mean, na.rm = TRUE)
agg_mult <- aggregate(seasonal ~ month, data = df_mult, FUN = mean, na.rm = TRUE)
agg_stl  <- aggregate(seasonal ~ month, data = df_s,    FUN = mean, na.rm = TRUE)

mk_month <- function(df_avg, ttl, ref_line) {
  ggplot(df_avg, aes(month, seasonal)) +
    geom_hline(yintercept = ref_line, linetype = "dashed", size = 0.25, col='red') +
    geom_line(size = 0.45) +
    geom_point(size = 1) +
    scale_x_continuous(
      breaks = 1:12,
      labels = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
    ) +
    labs(title = ttl, x = "Month", y = NULL) +
    base_theme +
    theme(plot.title = element_text(size = 9))
}

q_add  <- mk_month(agg_add,  "Average seasonal profile : Additive", 0)
q_stl  <- mk_month(agg_stl,  "Average seasonal profile : STL", 0)
q_mult <- mk_month(agg_mult, "Average seasonal profile : Multiplicative", 1)

gridExtra::grid.arrange(q_add, q_stl, q_mult, ncol = 3)
```
The aggregated monthly profiles confirm a clear and regular annual seasonal cycle. For instance, late summer months such as August tend to exhibit negative seasonal effects, whereas autumn months such as October often contribute positively to the seasonal component. This produces a stable intra-annual structure that is remarkably consistent across additive, multiplicative, and STL decompositions. Although stl() allows for time-varying seasonality through LOESS smoothing, the close similarity of the results suggests that seasonality remains approximately constant over time, making the classical constant-seasonality assumption appropriate for the GISTEMP data.

```{r task1_residuals, fig.width=13, fig.height=4, message=FALSE, warning=FALSE}
library(ggplot2)
library(gridExtra)

mk_residuals <- function(df, ttl, ref_line) {
  ggplot(df, aes(t, residual)) +
    geom_line(size = 0.25) +
    geom_hline(yintercept = ref_line, linetype = "dashed", size = 0.25, col='red') +
    labs(title = ttl, x = "Year", y = NULL) +
    base_theme +
    theme(plot.title = element_text(size = 9))
}

res_add  <- mk_residuals(df_add,"Residuals : Additive",0)
res_stl  <- mk_residuals(df_s,"Residuals : STL",0)
res_mult <- mk_residuals(df_mult,"Residuals : Multiplicative",1)

gridExtra::grid.arrange(res_add, res_stl, res_mult, ncol = 3)
```
The residuals fluctuate around zero without displaying persistent patterns, which supports the adequacy of the decomposition. While occasional extreme spikes are visible, they remain isolated and do not suggest structural misspecification. Overall, the unexplained variation appears mostly irregular rather than systematic, and no strong evidence of remaining seasonality or trend is detected in the residuals.

### Task 2

```{r}
seasonal_component <- stl_fit$time.series[, "seasonal"]
deseasonalized <- gistemp_ts - seasonal_component
train <- window(deseasonalized, start = c(1950, 1), end = c(2010, 12))

par(mar = c(4,4,2,1))  # cleaner margins

plot(train,
     main = "Deseasonalized Northern Hemisphere temperature anomalies (1950–2010)",
     xlab = "Year",
     ylab = "Temperature anomaly (°C)",
     col = "black",
     lwd = 1,
     cex.main = 1,
     cex.lab = 0.9,
     cex.axis = 0.8)

abline(h = 0, lty = 2, col = "red")

```

```{r}

library(forecast)

## --- Fit Holt's linear method (beta = TRUE, gamma = FALSE) ---
fit_holt <- holt(train, h = 1)

alpha_holt <- fit_holt$model$par["alpha"]
beta_holt  <- fit_holt$model$par["beta"]

alpha_holt  # 0.288203
beta_holt   # 0.0003834748

## --- One-step-ahead fitted values vs observed ---
par(mar = c(4, 4, 2, 1))

plot(train,
     main     = "Holt one-step-ahead forecasts vs observed (1950–2010)",
     xlab     = "Year",
     ylab     = "Temperature anomaly (°C)",
     col      = "black",
     lwd      = 1,
     cex.main = 1,
     cex.lab  = 0.9,
     cex.axis = 0.8)

lines(fitted(fit_holt), col = "red", lwd = 1.2)

legend("topleft",
       legend = c("Observed", "One-step-ahead forecast"),
       col    = c("black", "red"),
       lwd    = c(1, 1.2),
       cex    = 0.8,
       bty    = "n")
```

```{r}
## --- Holt-Winters with trend (beta = TRUE, gamma = FALSE) ---
fit_holt2 <- HoltWinters(train, beta = TRUE, gamma = FALSE)

# Extract parameters correctly
alpha_holt2 <- fit_holt2$alpha   # numeric estimated alpha
beta_holt2  <- fit_holt2$beta    # numeric estimated beta — this is the fix!

alpha_holt2
beta_holt2

## --- One-step-ahead fitted values vs observed ---
par(mar = c(4, 4, 2, 1))

plot(train,
     main    = "Holt-Winters one-step-ahead forecasts vs observed (1950–2010)",
     xlab    = "Year",
     ylab    = "Temperature anomaly (°C)",
     col     = "black",
     lwd     = 1,
     cex.main = 1,
     cex.lab  = 0.9,
     cex.axis = 0.8)

lines(fitted(fit_holt2)[, "xhat"], col = "red", lwd = 1.2)

legend("topleft",
       legend = c("Observed", "One-step-ahead forecast"),
       col    = c("black", "red"),
       lwd    = c(1, 1.2),
       cex    = 0.8,
       bty    = "n")

```

```{r}
# Forecasting the deseasonalized series (15 years ahead = 180 months)
library(forecast)

h_years <- 15
h_months <- h_years * 12

# --- Refit/compute forecasts with holt() (forecast package) on the deseasonalized train ---
# (train is already the deseasonalized ts for Jan1950-Dec2010 from your previous code)
holt_fc <- holt(train, h = h_months, level = c(80, 95), damped = FALSE)

# -- Forecast from your HoltWinters fit (already named fit_holt2) --
# fit_holt2 was fitted to 'train' earlier via HoltWinters(train, beta=TRUE, gamma=FALSE)
hw_fc <- forecast(fit_holt2, h = h_months, level = c(80, 95))

# --- Plot deseasonalized observed (train) and forecasts (both methods) ---
par(mfrow = c(1,2), mar = c(4,4,2,1))

# Panel A: holt() forecast (deseasonalized)
plot(window(deseasonalized, start = c(1950,1), end = c(2010,12)),
     main = "Holt forecasts on deseasonalized series (1950–2010 train)",
     xlab = "Year", ylab = "Deseasonalized anomaly (°C)",
     col = "black", lwd = 1)
# add holt point forecasts and intervals
lines(holt_fc$mean, col = "red", lwd = 1.2)
# 80% interval
lines(ts(holt_fc$lower[ ,1], start = time(holt_fc$mean)[1], frequency = 12), col = "red", lty = 2)
lines(ts(holt_fc$upper[ ,1], start = time(holt_fc$mean)[1], frequency = 12), col = "red", lty = 2)
# 95% interval
lines(ts(holt_fc$lower[ ,2], start = time(holt_fc$mean)[1], frequency = 12), col = "red", lty = 3)
lines(ts(holt_fc$upper[ ,2], start = time(holt_fc$mean)[1], frequency = 12), col = "red", lty = 3)

legend("topleft", legend = c("Observed (train)", "Holt mean", "80% PI", "95% PI"),
       col = c("black", "red", "red", "red"), lwd = c(1,1.2,1,1), lty = c(1,1,2,3),
       bty = "n", cex = 0.8)

# Panel B: Holt-Winters (HoltWinters fitted to train) forecast
plot(window(deseasonalized, start = c(1950,1), end = c(2010,12)),
     main = "Holt-Winters (via forecast()) on deseasonalized series",
     xlab = "Year", ylab = "Deseasonalized anomaly (°C)",
     col = "black", lwd = 1)
lines(hw_fc$mean, col = "blue", lwd = 1.2)
lines(ts(hw_fc$lower[ ,1], start = time(hw_fc$mean)[1], frequency = 12), col = "blue", lty = 2)
lines(ts(hw_fc$upper[ ,1], start = time(hw_fc$mean)[1], frequency = 12), col = "blue", lty = 2)
lines(ts(hw_fc$lower[ ,2], start = time(hw_fc$mean)[1], frequency = 12), col = "blue", lty = 3)
lines(ts(hw_fc$upper[ ,2], start = time(hw_fc$mean)[1], frequency = 12), col = "blue", lty = 3)

legend("topleft", legend = c("Observed (train)", "HW mean", "80% PI", "95% PI"),
       col = c("black", "blue", "blue", "blue"), lwd = c(1,1.2,1,1), lty = c(1,1,2,3),
       bty = "n", cex = 0.8)

# reset plotting
par(mfrow = c(1,1))

# --- MAE evaluation at horizons if observed out-of-sample exists (deseasonalized) ---
# observed out-of-sample deseasonalized values (2011-...): may be shorter than full 15 years depending on your CSV
obs_oos <- window(deseasonalized, start = c(2011,1), end = c(2010 + h_years, 12))

horizons <- c(1, 5, 10, 15)
mae_table <- data.frame(horizon_years = horizons, MAE_holt = NA_real_, MAE_hw = NA_real_)

for (i in seq_along(horizons)) {
  yrs <- horizons[i]
  m <- yrs * 12
  # ensure we have at least m observed months after 2010
  if (length(obs_oos) >= m) {
    obs_slice <- as.numeric(window(deseasonalized, start = c(2011,1), end = c(2010 + yrs, 12)))
    holt_slice <- as.numeric(holt_fc$mean[1:m])
    hw_slice   <- as.numeric(hw_fc$mean[1:m])
    mae_table$MAE_holt[i] <- mean(abs(obs_slice - holt_slice))
    mae_table$MAE_hw[i]   <- mean(abs(obs_slice - hw_slice))
  } else {
    mae_table$MAE_holt[i] <- NA
    mae_table$MAE_hw[i]   <- NA
  }
}

# write MAE table for the report (and print to console for quick check)
write.csv(mae_table, "mae_task2_holt_vs_hw.csv", row.names = FALSE)
mae_table
```

```{r}
## --- 15-year forecast from end of 2010 ---
fit_holt_forecast <- holt(train, h = 15 * 12)

autoplot(fit_holt_forecast) +
  autolayer(train, series = "Observed") +
  labs(
    title  = "15-year Holt forecast of deseasonalized temperature anomalies (2011–2025)",
    x      = "Year",
    y      = "Temperature anomaly (°C)"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.title = element_blank())
```


```{r}
library(forecast)
library(ggplot2)

## --- 15-year forecast from end of 2010 ---
fit_holt_forecast <- forecast(holt(train, h = 15 * 12))

fc_df <- data.frame(
  time = as.numeric(time(fit_holt_forecast$mean)),
  mean = as.numeric(fit_holt_forecast$mean),
  lo80 = as.numeric(fit_holt_forecast$lower[, 1]),
  hi80 = as.numeric(fit_holt_forecast$upper[, 1]),
  lo95 = as.numeric(fit_holt_forecast$lower[, 2]),
  hi95 = as.numeric(fit_holt_forecast$upper[, 2])
)

train_df <- data.frame(
  time  = as.numeric(time(train)),
  value = as.numeric(train)
)

ggplot() +
  geom_ribbon(data = fc_df, aes(x = time, ymin = lo95, ymax = hi95), fill = "steelblue", alpha = 0.2) +
  geom_ribbon(data = fc_df, aes(x = time, ymin = lo80, ymax = hi80), fill = "steelblue", alpha = 0.3) +
  geom_line(data = train_df, aes(x = time, y = value), colour = "black", linewidth = 0.5) +
  geom_line(data = fc_df, aes(x = time, y = mean), colour = "steelblue", linewidth = 0.8) +
  labs(
    title = "15-year Holt forecast of deseasonalized temperature anomalies (2011–2025)",
    x     = "Year",
    y     = "Temperature anomaly (°C)"
  ) +
  theme_minimal(base_size = 10)
```
```{r}
## --- MAE at horizons: 1, 5, 10, 15 years ---
actual <- window(deseasonalized, start = c(2011, 1), end = c(2025, 12))
forecasts <- fit_holt_forecast$mean  # point forecasts from holt

horizons <- c(1, 5, 10, 15)
mae_results <- sapply(horizons, function(yr) {
  h <- yr * 12
  mean(abs(forecasts[1:h] - actual[1:h]))
})

mae_table <- data.frame(
  horizon_years = horizons,
  MAE_holt      = mae_results
)

knitr::kable(mae_table, digits = 7, align = c("r", "r"),
             caption = "MAE at increasing forecast horizons")

>>>>>>> a6d058d (Task 1 final version)
```

```{r}
## --- Holt-Winters with trend (beta = TRUE, gamma = FALSE) ---
fit_holt2 <- HoltWinters(train, beta = TRUE, gamma = FALSE)

# Extract parameters correctly
alpha_holt2 <- fit_holt2$alpha   # numeric estimated alpha
beta_holt2  <- fit_holt2$beta    # numeric estimated beta — this is the fix!

alpha_holt2
beta_holt2

## --- One-step-ahead fitted values vs observed ---
par(mar = c(4, 4, 2, 1))

plot(train,
     main    = "Holt-Winters one-step-ahead forecasts vs observed (1950–2010)",
     xlab    = "Year",
     ylab    = "Temperature anomaly (°C)",
     col     = "black",
     lwd     = 1,
     cex.main = 1,
     cex.lab  = 0.9,
     cex.axis = 0.8)

lines(fitted(fit_holt2)[, "xhat"], col = "red", lwd = 1.2)

legend("topleft",
       legend = c("Observed", "One-step-ahead forecast"),
       col    = c("black", "red"),
       lwd    = c(1, 1.2),
       cex    = 0.8,
       bty    = "n")

```

```{r}
# Forecasting the deseasonalized series (15 years ahead = 180 months)
library(forecast)

h_years <- 15
h_months <- h_years * 12

# --- Refit/compute forecasts with holt() (forecast package) on the deseasonalized train ---
# (train is already the deseasonalized ts for Jan1950-Dec2010 from your previous code)
holt_fc <- holt(train, h = h_months, level = c(80, 95), damped = FALSE)

# -- Forecast from your HoltWinters fit (already named fit_holt2) --
# fit_holt2 was fitted to 'train' earlier via HoltWinters(train, beta=TRUE, gamma=FALSE)
hw_fc <- forecast(fit_holt2, h = h_months, level = c(80, 95))

# --- Plot deseasonalized observed (train) and forecasts (both methods) ---
par(mfrow = c(1,2), mar = c(4,4,2,1))

# Panel A: holt() forecast (deseasonalized)
plot(window(deseasonalized, start = c(1950,1), end = c(2010,12)),
     main = "Holt forecasts on deseasonalized series (1950–2010 train)",
     xlab = "Year", ylab = "Deseasonalized anomaly (°C)",
     col = "black", lwd = 1)
# add holt point forecasts and intervals
lines(holt_fc$mean, col = "red", lwd = 1.2)
# 80% interval
lines(ts(holt_fc$lower[ ,1], start = time(holt_fc$mean)[1], frequency = 12), col = "red", lty = 2)
lines(ts(holt_fc$upper[ ,1], start = time(holt_fc$mean)[1], frequency = 12), col = "red", lty = 2)
# 95% interval
lines(ts(holt_fc$lower[ ,2], start = time(holt_fc$mean)[1], frequency = 12), col = "red", lty = 3)
lines(ts(holt_fc$upper[ ,2], start = time(holt_fc$mean)[1], frequency = 12), col = "red", lty = 3)

legend("topleft", legend = c("Observed (train)", "Holt mean", "80% PI", "95% PI"),
       col = c("black", "red", "red", "red"), lwd = c(1,1.2,1,1), lty = c(1,1,2,3),
       bty = "n", cex = 0.8)

# Panel B: Holt-Winters (HoltWinters fitted to train) forecast
plot(window(deseasonalized, start = c(1950,1), end = c(2010,12)),
     main = "Holt-Winters (via forecast()) on deseasonalized series",
     xlab = "Year", ylab = "Deseasonalized anomaly (°C)",
     col = "black", lwd = 1)
lines(hw_fc$mean, col = "blue", lwd = 1.2)
lines(ts(hw_fc$lower[ ,1], start = time(hw_fc$mean)[1], frequency = 12), col = "blue", lty = 2)
lines(ts(hw_fc$upper[ ,1], start = time(hw_fc$mean)[1], frequency = 12), col = "blue", lty = 2)
lines(ts(hw_fc$lower[ ,2], start = time(hw_fc$mean)[1], frequency = 12), col = "blue", lty = 3)
lines(ts(hw_fc$upper[ ,2], start = time(hw_fc$mean)[1], frequency = 12), col = "blue", lty = 3)

legend("topleft", legend = c("Observed (train)", "HW mean", "80% PI", "95% PI"),
       col = c("black", "blue", "blue", "blue"), lwd = c(1,1.2,1,1), lty = c(1,1,2,3),
       bty = "n", cex = 0.8)

# reset plotting
par(mfrow = c(1,1))

# --- MAE evaluation at horizons if observed out-of-sample exists (deseasonalized) ---
# observed out-of-sample deseasonalized values (2011-...): may be shorter than full 15 years depending on your CSV
obs_oos <- window(deseasonalized, start = c(2011,1), end = c(2010 + h_years, 12))

horizons <- c(1, 5, 10, 15)
mae_table <- data.frame(horizon_years = horizons, MAE_holt = NA_real_, MAE_hw = NA_real_)

for (i in seq_along(horizons)) {
  yrs <- horizons[i]
  m <- yrs * 12
  # ensure we have at least m observed months after 2010
  if (length(obs_oos) >= m) {
    obs_slice <- as.numeric(window(deseasonalized, start = c(2011,1), end = c(2010 + yrs, 12)))
    holt_slice <- as.numeric(holt_fc$mean[1:m])
    hw_slice   <- as.numeric(hw_fc$mean[1:m])
    mae_table$MAE_holt[i] <- mean(abs(obs_slice - holt_slice))
    mae_table$MAE_hw[i]   <- mean(abs(obs_slice - hw_slice))
  } else {
    mae_table$MAE_holt[i] <- NA
    mae_table$MAE_hw[i]   <- NA
  }
}

# write MAE table for the report (and print to console for quick check)
write.csv(mae_table, "mae_task2_holt_vs_hw.csv", row.names = FALSE)
mae_table
```

```{r}
## --- 15-year forecast from end of 2010 ---
fit_holt_forecast <- holt(train, h = 15 * 12)

autoplot(fit_holt_forecast) +
  autolayer(train, series = "Observed") +
  labs(
    title  = "15-year Holt forecast of deseasonalized temperature anomalies (2011–2025)",
    x      = "Year",
    y      = "Temperature anomaly (°C)"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.title = element_blank())
```


```{r}
library(forecast)
library(ggplot2)

## --- 15-year forecast from end of 2010 ---
fit_holt_forecast <- forecast(holt(train, h = 15 * 12))

fc_df <- data.frame(
  time = as.numeric(time(fit_holt_forecast$mean)),
  mean = as.numeric(fit_holt_forecast$mean),
  lo80 = as.numeric(fit_holt_forecast$lower[, 1]),
  hi80 = as.numeric(fit_holt_forecast$upper[, 1]),
  lo95 = as.numeric(fit_holt_forecast$lower[, 2]),
  hi95 = as.numeric(fit_holt_forecast$upper[, 2])
)

train_df <- data.frame(
  time  = as.numeric(time(train)),
  value = as.numeric(train)
)

ggplot() +
  geom_ribbon(data = fc_df, aes(x = time, ymin = lo95, ymax = hi95), fill = "steelblue", alpha = 0.2) +
  geom_ribbon(data = fc_df, aes(x = time, ymin = lo80, ymax = hi80), fill = "steelblue", alpha = 0.3) +
  geom_line(data = train_df, aes(x = time, y = value), colour = "black", linewidth = 0.5) +
  geom_line(data = fc_df, aes(x = time, y = mean), colour = "steelblue", linewidth = 0.8) +
  labs(
    title = "15-year Holt forecast of deseasonalized temperature anomalies (2011–2025)",
    x     = "Year",
    y     = "Temperature anomaly (°C)"
  ) +
  theme_minimal(base_size = 10)
```

```{r}
## --- MAE at horizons: 1, 5, 10, 15 years ---
actual <- window(deseasonalized, start = c(2011, 1), end = c(2025, 12))
forecasts <- fit_holt_forecast$mean  # point forecasts from holt

horizons <- c(1, 5, 10, 15)
mae_results <- sapply(horizons, function(yr) {
  h <- yr * 12
  mean(abs(forecasts[1:h] - actual[1:h]))
})

mae_table <- data.frame(
  horizon_years = horizons,
  MAE_holt      = mae_results
)

knitr::kable(mae_table, digits = 7, align = c("r", "r"),
             caption = "MAE at increasing forecast horizons")

```

