---
title: "Bocconi University - 20236 Time Series Analysis"
subtitle: "**Assignment 1 towards the final project | Group 27**"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: yes
    extra_dependencies: ["float"]
fontsize: 10pt
geometry: margin=2cm
---

```{r setup, include=FALSE}
rm(list = ls())
library(forecast)
library(tidyverse)
library(gridExtra)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center", fig.pos = "H", out.extra = "")

set_theme(theme_classic())
update_theme(plot.title = element_text(hjust = 0.5), strip.background = element_blank())

plot_ts_gg <- function(
    x,
    title = NULL,
    ylab = NULL,
    aes = NULL,
    facet_ncol = 1,
    wrap_width = 45
) {
  if (!inherits(x, "ts")) {
    stop("Input must be ts.")
  }

  df <- tibble(
    Time = as.numeric(time(x)),
    Value = as.numeric(x)
  )

  p <- ggplot(df, aes(Time, Value, !!!aes)) +
    geom_line() +
    labs(title = title, x = NULL, y = ylab)

  p
}
```

# Exploratory Analysis

```{r, fig.height=3, fig.width=5, fig.height = 3, fig.show = "hold", out.width = "50%", fig.cap = "\\label{fig:main}Monthly Northern Hemisphere temperature anomalies (GISTEMP)."}
df <- read.csv("gistemp.csv", skip = 1, na.strings = "***")
#head(df)
monthly_matrix <- as.matrix(df[, 2:13])
monthly_vector <- as.numeric(t(monthly_matrix))

gistemp_ts <- ts(monthly_vector, start = c(1880, 1), frequency = 12)

plot_ts_gg(
  gistemp_ts,
  title = "Temperature anomalies (°C)"
) +
  geom_hline(yintercept = 0)


```

The time series (Figure$~$\ref{fig:main}) shows a long-term upward trend in temperature anomalies over the sample period. Values are mostly negative before the mid-20th century and increase gradually over time. From approximately the 1980s onward, the rise becomes more pronounced, suggesting a possible acceleration in warming. Short-term fluctuations remain visible around this trend.

The marginal variances appear relatively stable, as the dispersion does not increase with the level of the series. There is no strong visual evidence of heteroskedasticity. The variance can therefore be considered approximately constant over time. This observation supports the use of an additive decomposition rather than a multiplicative one.

## Task 1

```{r task1_decompo_prep, message=FALSE, warning=FALSE}
# Ensuring positive values, necessary for classical multiplicative decomposition
shift <- 0
if (min(gistemp_ts, na.rm = TRUE) <= 0) {
  shift <- abs(min(gistemp_ts, na.rm = TRUE)) + 0.01
}
gistemp_pos <- gistemp_ts + shift
```


```{r task1_decompo, fig.height=3, fig.width=5, fig.show = "hold", out.width = "50%", fig.cap = "\\label{fig:decomp}Comparison of time-series decompositions of Northern Hemisphere temperature anomalies."}
dec_add  <- decompose(gistemp_ts,  type = "additive")
dec_mult <- decompose(gistemp_pos, type = "multiplicative")
stl_fit  <- stl(gistemp_ts, s.window = "periodic")

# Assuming same seasonality for all decompositions
const_seasonality <- frequency(gistemp_ts)

# Conversion of the decompose() object for a cleaner output
df_decomp <- function(dec_obj) {
  data.frame(
    t = as.numeric(time(dec_obj$x)),
    observed = as.numeric(dec_obj$x),
    trend = as.numeric(dec_obj$trend),
    seasonal = as.numeric(dec_obj$seasonal),
    residual = as.numeric(dec_obj$random)
  )
}

df_stl <- function(stl_obj, y) {
  t <- as.numeric(time(stl_obj$time.series[, 1]))
  data.frame(
    t = t,
    observed = as.numeric(y),
    trend = as.numeric(stl_obj$time.series[, "trend"]),
    seasonal = as.numeric(stl_obj$time.series[, "seasonal"]),
    residual = as.numeric(stl_obj$time.series[, "remainder"])
  )
}

df_hw <- function(hw_obj, y, pad = 2) {
  t <- as.numeric(time(y))
  tibble(
    t = t,
    observed = as.numeric(y),
    fitted = c(rep(NA, 2), as.numeric(fitted(hw_obj)[, "xhat"])),
    trend = c(rep(NA, 2), as.numeric(fitted(hw_obj)[, "level"]) + as.numeric(fitted(hw_obj)[, "trend"])),
    residual = observed - fitted
  )
}

# Formatting as a panel using subplots
panel <- function(df, main_title, add_hline = TRUE) {
  p_observed <- ggplot(df, aes(t, observed)) + geom_line(size = 0.3) +
    labs(title = "Observed", x = NULL, y = NULL)
  if (add_hline) {
    p_observed <- p_observed + geom_hline(yintercept = 0)
  }
  p_trend <- ggplot(df, aes(t, trend)) + geom_line(size = 0.3) +
    labs(title = "Trend", x = NULL, y = NULL)

  seasonal_seq <- seq(from = 1, to = const_seasonality)

  p_seasonal <- df %>%
    slice_head(n = const_seasonality) %>%
    mutate(t = seasonal_seq) %>%
    ggplot(aes(t, seasonal)) + geom_line(size = 0.3) +
    scale_x_continuous(breaks = seasonal_seq) +
    labs(title = "Seasonal", x = NULL, y = NULL)


  p_residual <- ggplot(df, aes(t, residual)) + geom_line(size = 0.3) +
    labs(title = "Residual", x = NULL, y = NULL)

  # Choosing ncol = 2
  gridExtra::grid.arrange(
    p_observed, p_trend,
    p_seasonal, p_residual,
    ncol = 2,
    top = grid::textGrob(main_title, gp = grid::gpar(fontface = "bold", fontsize = 14))
  )
}

# Using grid to display panels
grid::grid.draw(panel(df_decomp(dec_add),  "Additive decomposition (GISTEMP)"))
grid::grid.draw(panel(df_decomp(dec_mult), "Multiplicative decomposition (shifted GISTEMP)", add_hline = FALSE))
grid::grid.draw(panel(df_stl(stl_fit, gistemp_ts), "STL decomposition (periodic seasonality)"))

```
bitte fügen Sie hier den kommentar ein
Answer the following questions, supporting your answers with valid analysis when necessary
(at your discretion).
Trend Analysis:
• What long-term trends are visible in the temperature data?
• Are there periods of acceleration or deceleration in warming?
Seasonality:
• Can you interpret the seasonal pattern?
Is the seasonality constant, or time–varying? Explain how different R functions tackle
the problem; explore both constant and time-varying cases and pick the one you reckon
is the most suitable for the data.
Noise and Uncertainty:
• Briefly comment on the variation that is unexplained by the decomposition. Are there
anomalies in the residuals?

## Task 2

### SES or Holt-Winters

fig.height=3, fig.width=5, fig.show = "hold", out.width = "50%"}

```{r, fig.height = 3, fig.width = 5, fig.show = "hold", out.width = "50%", fig.cap = "\\label{fig:hw}Deseasonalized Northern Hemisphere temperature anomalies (1950–2010) and Holt–Winters one-step ahead forceasts."}
seasonal_component <- stl_fit$time.series[, "seasonal"]
deseasonalized <- gistemp_ts - seasonal_component
train <- window(deseasonalized, start = c(1950, 1), end = c(2010, 12))

## HW with trend (gamma = FALSE)
fit_holtwinters <- HoltWinters(train, gamma = FALSE, optim.start = c(alpha = 0.8, beta = 0.9))

# Parameters extraction
alpha_holtwinters <- fit_holtwinters$alpha
beta_holtwinters  <- fit_holtwinters$beta

alpha_holtwinters
beta_holtwinters

plot_ts_gg(train,
  ylab = "Temperature anomaly (°C)",
  aes = list(color = "Observed"),
) +
  geom_hline(yintercept = 0) +
  geom_line(data = df_hw(fit_holtwinters, train), aes(x = t, y = fitted, color = "One-step-ahead forecast")) +
  scale_color_manual(values = c("black", "#0072B2")) +
  theme(legend.position = "inside", legend.position.inside = c(0.2, 0.9)) +
  labs(color = NULL)
```

What value of α have you used? Report it and explain how it is determined. How
would your forecasts change with a different choice of α?



### Forecasting the series

```{r, fig.height = 3, fig.width = 5, fig.show = "hold", out.width = "50%", fig.cap = "\\label{fig:hw_pred}15-year Holt–Winters forecast of deseasonalized temperature anomalies (2011–2025) with 80% and 95% confidence bands."}
# 15-year forecast from end of 2010
fit_holtwinters_forecast <- forecast(fit_holtwinters, h = 15 * 12)

fc_df <- data.frame(
  time = as.numeric(time(fit_holtwinters_forecast$mean)),
  mean = as.numeric(fit_holtwinters_forecast$mean),
  lo80 = as.numeric(fit_holtwinters_forecast$lower[, 1]),
  hi80 = as.numeric(fit_holtwinters_forecast$upper[, 1]),
  lo95 = as.numeric(fit_holtwinters_forecast$lower[, 2]),
  hi95 = as.numeric(fit_holtwinters_forecast$upper[, 2])
)

train_df <- data.frame(
  time  = as.numeric(time(train)),
  value = as.numeric(train)
)

ggplot() +
  geom_ribbon(data = fc_df, aes(x = time, ymin = lo95, ymax = hi95), fill = "steelblue", alpha = 0.2) +
  geom_ribbon(data = fc_df, aes(x = time, ymin = lo80, ymax = hi80), fill = "steelblue", alpha = 0.3) +
  geom_line(data = train_df, aes(x = time, y = value), colour = "black", linewidth = 0.5) +
  geom_line(data = fc_df, aes(x = time, y = mean), colour = "steelblue", linewidth = 0.8) +
  labs(
    x     = "Year",
    y     = "Temperature anomaly (°C)"
  ) +
  geom_hline(yintercept = 0)
```
```{r}
# MAE at horizons: 1, 5, 10, 15 years
actual <- window(deseasonalized, start = c(2011, 1), end = c(2025, 12))
forecasts <- fit_holtwinters_forecast$mean  # point forecasts from holt

horizons <- c(1, 5, 10, 15)
mae_results <- sapply(horizons, function(yr) {
  h <- yr * 12
  mean(abs(forecasts[1:h] - actual[1:h]))
})

mae_table <- data.frame(
  horizon_years = horizons,
  MAE_holtwinters = mae_results
)

knitr::kable(mae_table, digits = 7, align = c("r", "r"),
             caption = "MAE at increasing forecast horizons")

```

Briefly discuss your results. Do you envisage any pattern?










