---
title: "Bocconi University - 20236 Time Series Analysis"
subtitle: "**Group 27 | Anna-Bella Dreyfus, Emanuele Artoni, Artūrs Jānis Ņikitins,
  Madelief van Weerdenburg**"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
    extra_dependencies: float
  html_document:
    df_print: paged
fontsize: 10pt
geometry: margin=2cm
header-includes:
- \setlength{\textfloatsep}{6pt}
- \setlength{\intextsep}{6pt}
---

```{r setup, include=FALSE}
rm(list = ls())
library(depmixS4)
library(tidyverse)
library(gridExtra)
library(stargazer)
library(ggplot2)
library(forecast)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center", fig.pos = "H", out.extra = "")

set_theme(theme_classic())
update_theme(plot.title = element_text(hjust = 0.5), strip.background = element_blank())

```

```{r implementation}
df <- read.csv("gistemp.csv", skip = 1, na.strings = "***")
monthly_matrix <- as.matrix(df[, 2:13])
monthly_vector <- as.numeric(t(monthly_matrix))

gistemp_ts <- ts(monthly_vector, start = c(1880, 1), frequency = 12)
gistemp_stl <- stl(gistemp_ts, s.window = "periodic")
gistemp_adj <- seasadj(gistemp_stl)
```
### Choice of a Hidden Markov Model

In the context of GISTEMP temperature data, the dynamics are likely non-stationary, with possible structural changes such as periods of stable temperatures followed by phases of accelerated warming. Standard time series models may fail to capture these features, whereas Hidden Markov Models allow the data to depend on an underlying latent state that can switch over time, making them well suited to identify such regime changes.

###State-dependent level

We first consider Gaussian Hidden Markov Models with state-dependent means and variances. To assess the appropriate number of latent regimes, we estimate both a 2-state and a 3-state specification, and compare them using log-likelihood, AIC and BIC.

```{r hmm_level}
library(depmixS4)
library(tidyr)
library(dplyr)
library(ggplot2)

data_hmm <- data.frame(
  time_index = as.numeric(time(gistemp_adj)),
  y = as.numeric(gistemp_adj)
)
set.seed(555)
#2-state Gaussian HMM
mod_hmm_2 <- depmix(y ~ 1, data = data_hmm, nstates = 2)
fit_hmm_2 <- fit(mod_hmm_2)

#3-state Gaussian HMM
mod_hmm_3 <- depmix(y ~ 1, data = data_hmm, nstates = 3)
fit_hmm_3 <- fit(mod_hmm_3)

comparison_table <- data.frame(
  Model = c("2-state HMM", "3-state HMM"),
  logLik = c(logLik(fit_hmm_2), logLik(fit_hmm_3)),
  AIC = c(AIC(fit_hmm_2), AIC(fit_hmm_3)),
  BIC = c(BIC(fit_hmm_2), BIC(fit_hmm_3))
)

comparison_table
summary(fit_hmm_2)
summary(fit_hmm_3)
```
The 3-state model provides a substantially better fit, with a higher log-likelihood and lower AIC and BIC than the 2-state model. This indicates that allowing for an additional regime improves the representation of the temperature dynamics.

Indeed, while the 2-state model captures a simple contrast between low and high temperature levels, the 3-state model identifies a more nuanced structure, distinguishing between low, intermediate and high regimes. This appears more consistent with the gradual evolution of climate patterns.

Therefore, the 3-state model is preferred as it improves the fit. However, its validity should be further assessed by examining the stability and interpretability of the decoded regimes, to rule out potential overfitting.






```{r decoding}
states_3 <- posterior(fit_hmm_3)

data_hmm$state_3 <- factor(states_3$state)

ggplot(data_hmm, aes(x = time_index, y = y)) +
  geom_line(color = "black", linewidth = 0.35) +
  geom_point(aes(color = state_3), size = 0.5, alpha = 0.7) +
  labs(
    title = "Decoded regimes from the 3-state HMM",
    x = "Year",
    y = "Temperature anomaly",
    color = "Regime"
  ) +
  theme_minimal()
```
The decoded regimes appear stable and clearly interpretable over time. Transitions between states are relatively infrequent and follow a coherent long-term pattern, suggesting that the model captures meaningful structural changes rather than overfitting.

```{r}

# data_hmm_new <- data_hmm[data_hmm$time_index < 1976, ]


# set.seed(555)
# nmod_1 <- depmix(y ~ 1, data = data_hmm, nstates = 2)
# f_nmod_1 <- fit(nmod_1)
# est_States_1 <- posterior(f_nmod_1)

# nmod_2 <- depmix(y ~ 1, data = data_hmm, nstates = 3)
# f_nmod_2 <- fit(nmod_2)
# est_States_2 <- posterior(f_nmod_2)

# nmod_3 <- depmix(y ~ 1, data = data_hmm, nstates = 4)
# f_nmod_3 <- fit(nmod_3)
# est_States_3 <- posterior(f_nmod_3)

# comparison_table <- data.frame(
#   Model = c("2-state HMM", "3-state HMM", "test"),
#   logLik = c(logLik(f_nmod_1), logLik(f_nmod_2), logLik(f_nmod_3)),
#   AIC = c(AIC(f_nmod_1), AIC(f_nmod_2), AIC(f_nmod_3)),
#   BIC = c(BIC(f_nmod_1), BIC(f_nmod_2), BIC(f_nmod_3))
# )

# f_nmod_1@response[[1]][[1]]@parameters$coefficients

# # computing est_mean trajectory for subsample model 1 (2 states)

# testy <- lapply(1:2, function(i) {
#   lapply(f_nmod_1@response, function(resp) {
#     resp[[1]]@parameters$coefficients
#   })
# })

# testy


# sapply(1:3, function(i) {
#   f_nmod_2@response[[i]][[1]]@parameters$coefficients
# })


# sapply(1:4, function(i) {
#   f_nmod_3@response[[i]][[1]]@parameters$coefficients
# })




# # computing est_mean trajectory for subsample model 2 (3 states)
# est_mean_nmod2_state1 <- f_nmod_2@response[[1]][[1]]@parameters$coefficients
# est_mean_nmod2_state2 <- f_nmod_2@response[[2]][[1]]@parameters$coefficients
# est_mean_nmod2_state3 <- f_nmod_2@response[[3]][[1]]@parameters$coefficients

# est_mean_trajectory_nmod2 <- rep(est_mean_nmod2_state1, times = length(data_hmm_new$y))
# est_mean_trajectory_nmod2[est_States_2[,1] == 2] = est_mean_nmod2_state2
# est_mean_trajectory_nmod2[est_States_2[,1] == 3] = est_mean_nmod2_state3


# # two
# vis <- tibble(
#   time_index = abs(data_hmm_new$time_index),
#   estimated_mean = est_mean_trajectory_nmod2,
#   sample_trajectory = data_hmm_new$y
# ) %>%
#   tidyr::pivot_longer(
#     cols = c(estimated_mean, sample_trajectory),
#     names_to = "variable",
#     values_to = "Value"
#   )

# ggplot(vis, aes(x = time_index, y = Value)) +
#   geom_line() +
#   facet_wrap(~ variable, scales = "free", ncol = 1)


```

```{r multi-state fit}

fit_models <- \(data, max_states, reps, trend = FALSE, seed = 555) {
  set.seed(seed)

  models <- tibble()

  if (trend) {
    formula <- y ~ time_index
  } else {
    formula <- y ~ 1
  }

  for (n in 2:max_states) {
    models_iter <- tibble(iter = 1:reps, n = n) %>%
      rowwise() %>%
      mutate(mod = list(fit(depmix(formula, data = data, nstates = n))))

    models <- bind_rows(models, models_iter)
  }

  models <- models %>%
    rowwise() %>%
    mutate(
      across(mod, list(
        Model = ~ paste0(.@nstates, "-state HMM"),
        logLik = ~ as.numeric(logLik(.)),
        AIC = AIC,
        BIC = BIC
      ))
    ) %>%
    ungroup() %>%
    slice_max(mod_logLik, n = 1)

  models
}

# comparison_table <- fit_models(data_hmm, max_states = 8, reps = 30)
# saveRDS(comparison_table, "work/models.RDS")
comparison_table <- readRDS("work/models.RDS")

comparison_table %>%
  pivot_longer(c(mod_logLik, mod_BIC, mod_AIC), names_to = "indicator", values_to = "val", names_prefix = "mod_") %>%
  ggplot(aes(x = n, y = val)) +
  geom_line() +
  facet_wrap(~ indicator)

# BIC decreases until 6 states
```

```{r multi-state decoding}
# Check state interpretability and stability

## Select max likelihood model for each state choice
mod_selected <- comparison_table %>%
  filter(between(n, 3, 6))

decoding_table <- mod_selected %>%
  reframe(posterior(mod[[1]], type = "viterbi"), t = 1:length(state), y = data_hmm$y, .by = c(iter, n))

decoding_table %>%
  ggplot(aes(x = t, y = y)) +
  geom_line(aes(y = state), alpha = 0.3) +
  geom_line(alpha = 0.3) +
  geom_point(aes(color = factor(state)), size = 0.5, alpha = 0.7) +
  facet_wrap(~ n) +
  guides(color=guide_legend(override.aes=list(size=8), title = "State"))

# Consequence of only modelling a level: states are splitting up the increasing trend

decoding_table %>%
  count(n, state, name = "count") %>%
  mutate(prop = count / sum(count) * 100, .by = n) %>%
  ggplot(aes(x = state, y = prop)) +
  geom_col() +
  facet_wrap(~ n)

walk(mod_selected$mod, summary)

# 6 is too much, 5 needlessly splits the initial level, 4 splits the increased level, could be better than 3
# The problem is that we only have a state-dependent level, we are not taking into account the changing trend.
```

```{r multi-state short}
# Repeated for the shorter sample
data_hmm_short <- data_hmm[data_hmm$time_index < 1976, ]

# Fit
# comparison_table_short <- fit_models(data_hmm_short, max_states = 8, reps = 30)
# saveRDS(comparison_table_short, "work/models_short.RDS")
comparison_table_short <- readRDS("work/models_short.RDS")

comparison_table_short %>%
  pivot_longer(c(mod_logLik, mod_BIC, mod_AIC), names_to = "indicator", values_to = "val", names_prefix = "mod_") %>%
  ggplot(aes(x = n, y = val)) +
  geom_line() +
  facet_wrap(~ indicator)

# Improvement plateaus already at 3,4 states

# Decoding

# Check state interpretability and stability

## Select max likelihood model for each state choice
mod_selected_short <- comparison_table_short %>%
  filter(between(n, 2, 4))

decoding_table_short <- mod_selected_short %>%
  reframe(posterior(mod[[1]], type = "viterbi"), t = 1:length(state), y = data_hmm_short$y, .by = c(iter, n))

decoding_table_short %>%
  ggplot(aes(x = t, y = y)) +
  geom_line(aes(y = state), alpha = 0.3) +
  geom_line(alpha = 0.3) +
  geom_point(aes(color = factor(state)), size = 0.5, alpha = 0.7) +
  facet_wrap(~ n) +
  guides(color=guide_legend(override.aes=list(size=8), title = "State"))

# Consequence of only modelling a level: states are splitting up the increasing trend

decoding_table_short %>%
  count(n, state, name = "count") %>%
  mutate(prop = count / sum(count) * 100, .by = n) %>%
  ggplot(aes(x = state, y = prop)) +
  geom_col() +
  facet_wrap(~ n)

walk(mod_selected_short$mod, summary)
```

```{r multi-sample trend}
# Fit
# comparison_table_trend <- fit_models(data_hmm, max_states = 8, reps = 30, trend = TRUE)
# saveRDS(comparison_table_trend, "work/models_trend.RDS")
comparison_table_trend <- readRDS("work/models_trend.RDS")

comparison_table_trend %>%
  pivot_longer(c(mod_logLik, mod_BIC, mod_AIC), names_to = "indicator", values_to = "val", names_prefix = "mod_") %>%
  ggplot(aes(x = n, y = val)) +
  geom_line() +
  facet_wrap(~ indicator)

# More interesting, BIC bottoms out at 5

# Decoding

# Check state interpretability and stability

## Select max likelihood model for each state choice
mod_selected_trend <- comparison_table_trend %>%
  filter(between(n, 2, 6))

decoding_table_trend <- mod_selected_trend %>%
  reframe(posterior(mod[[1]], type = "viterbi"), t = 1:length(state), y = data_hmm$y, .by = c(iter, n))

decoding_table_trend %>%
  ggplot(aes(x = t, y = y)) +
  geom_line(aes(y = state), alpha = 0.3) +
  geom_line(alpha = 0.3) +
  geom_point(aes(color = factor(state)), size = 0.5, alpha = 0.7) +
  facet_wrap(~ n) +
  guides(color=guide_legend(override.aes=list(size=8), title = "State"))

decoding_table_trend %>%
  count(n, state, name = "count") %>%
  mutate(prop = count / sum(count) * 100, .by = n) %>%
  ggplot(aes(x = state, y = prop)) +
  geom_col() +
  facet_wrap(~ n)

walk(mod_selected_trend$mod, summary)
```
