---
title: "Bocconi University - 20236 Time Series Analysis"
subtitle: "**Group 27 | Anna-Bella Dreyfus, Emanuele Artoni, Artūrs Jānis Ņikitins,
  Madelief van Weerdenburg**"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
    extra_dependencies:
      - float
      - subfig
    dev: cairo_pdf
  html_document:
    df_print: paged
fontsize: 10pt
geometry: margin=2cm
header-includes:
- \setlength{\textfloatsep}{6pt}
- \setlength{\intextsep}{6pt}
---

```{r setup, include=FALSE}
rm(list = ls())
library(depmixS4)
library(tidyverse)
library(gridExtra)
library(stargazer)
library(ggplot2)
library(forecast)

knitr::opts_chunk$set(dev = "cairo_pdf", echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center", fig.pos = "H", out.extra = "")

set_theme(theme_classic(base_size = 10))
update_theme(plot.title = element_text(hjust = 0.5), strip.background = element_blank())

```

```{r implementation}
df <- read.csv("gistemp.csv", skip = 1, na.strings = "***")
monthly_matrix <- as.matrix(df[, 2:13])
monthly_vector <- as.numeric(t(monthly_matrix))

gistemp_ts <- ts(monthly_vector, start = c(1880, 1), frequency = 12)
gistemp_stl <- stl(gistemp_ts, s.window = "periodic")
gistemp_adj <- seasadj(gistemp_stl)

data_hmm <- data.frame(
  time_index = as.numeric(time(gistemp_adj)),
  y = as.numeric(gistemp_adj)
)
```

# Choice of a Hidden Markov Model

In the context of GISTEMP temperature data, the dynamics are likely non-stationary, with possible structural changes such as periods of stable temperatures followed by phases of accelerated warming. Standard time series models may fail to capture these features, whereas Hidden Markov Models allow the data to depend on an underlying latent state that can switch over time, making them well suited to identify such regime changes.

# State-dependent level

We first consider Gaussian Hidden Markov Models with state-dependent means and variances. To assess the appropriate number of latent regimes, we estimate both a 2-state and a 3-state specification, and compare them using log-likelihood, AIC and BIC.

# State-dependent linear trend

# Discussion

```{r multi-state base, results = FALSE, fig.show = "hide"}
fit_models <- \(data, fit_var = "y", max_states, reps, trend = FALSE, seed = 555) {
  set.seed(seed)

  models <- tibble()

  if (trend) {
    formula <- as.formula(paste(fit_var, "~ time_index"))
  } else {
    formula <- as.formula(paste(fit_var, "~ 1"))
  }

  for (n in 2:max_states) {
    models_iter <- tibble(iter = 1:reps, n = n) %>%
      rowwise() %>%
      mutate(mod = list(fit(depmix(formula, data = data, nstates = n))))

    models <- bind_rows(models, models_iter)
  }

  models <- models %>%
    rowwise() %>%
    mutate(
      across(mod, list(
        Model = ~ paste0(.@nstates, "-state HMM"),
        logLik = ~ as.numeric(logLik(.)),
        AIC = AIC,
        BIC = BIC
      ))
    ) %>%
    ungroup() %>%
    slice_max(mod_logLik, n = 1, by = n)

  models
}

# comparison_table <- fit_models(data_hmm, max_states = 8, reps = 30)
# saveRDS(comparison_table, "work/models.RDS")
comparison_table <- readRDS("work/models.RDS")

plot_ic_base <- comparison_table %>%
  pivot_longer(c(mod_logLik, mod_BIC, mod_AIC), names_to = "indicator", values_to = "val", names_prefix = "mod_") %>%
  ggplot(aes(x = n, y = val, color = indicator)) +
  geom_line() +
  labs(y = NULL, x = "State") +
  theme(legend.position = "top", legend.direction = "horizontal")

plot_ic_base

# BIC decreases until 6 states

# Check state interpretability and stability

## Select max likelihood model for each state choice
mod_selected <- comparison_table %>%
  filter(between(n, 4, 6))

response_table <- mod_selected %>%
  rowwise() %>%
  reframe(n = n, state_clas = 1:n, var = "y", as_tibble(matrix(getpars(mod, which = "pars")[-(1:((n + 1)*n))], ncol = 2, byrow = TRUE, dimnames = list(rep(NA, n), c("intercept", "sd"))))) %>%
  mutate(ord = rank(intercept), .by = n)

decoding_table <- mod_selected %>%
  reframe(posterior(mod[[1]], type = "viterbi"), t = data_hmm$time_index, y = data_hmm$y, .by = c(iter, n)) %>%
  mutate(state = response_table[response_table$n == cur_group()$n,]$ord[state], .by = n)

plot_comp_base <- decoding_table %>%
  select(n, t, y, state) %>%
  mutate(state_clas = state) %>%
  pivot_longer(c(y), names_to = "var", values_to = "val") %>%
  ggplot(aes(x = t, y = val)) +
  # geom_line(alpha = 0.3) +
  geom_point(aes(color = factor(state_clas)), alpha = 0.55, size = 1.3, pch=16) +
  geom_hline(data = response_table, aes(yintercept = intercept, group = factor(ord)), linewidth = 0.83) +
  geom_hline(data = response_table, aes(yintercept = intercept, color = factor(ord)), linewidth = 0.8) +
  facet_wrap(~ n, labeller = as_labeller(\(n) paste(n, "states"))) +
  guides(color=guide_legend(override.aes=list(size=5), title = "State")) +
  scale_color_viridis_d() +
  labs(y = "Temperature anomalies (°C)", x = NULL)

plot_comp_base

# Consequence of only modelling a level: states are splitting up the increasing trend

decoding_table %>%
  count(n, state, name = "count") %>%
  mutate(prop = count / sum(count) * 100, .by = n) %>%
  ggplot(aes(x = state, y = prop)) +
  geom_col() +
  facet_wrap(~ n)

walk(mod_selected$mod, summary)

# 6 is too much, 5 needlessly splits the initial level, 4 splits the increased level, could be better than 3
# The problem is that we only have a state-dependent level, we are not taking into account the changing trend.
```

```{r multi-state short, results = FALSE, fig.show = "hide"}
# Repeated for the shorter sample
data_hmm_short <- data_hmm[data_hmm$time_index < 1976, ]

# Fit
# comparison_table_short <- fit_models(data_hmm_short, max_states = 8, reps = 30)
# saveRDS(comparison_table_short, "work/models_short.RDS")
comparison_table_short <- readRDS("work/models_short.RDS")

plot_ic_short <- comparison_table_short %>%
  pivot_longer(c(mod_logLik, mod_BIC, mod_AIC), names_to = "indicator", values_to = "val", names_prefix = "mod_") %>%
  ggplot(aes(x = n, y = val, color = indicator)) +
  geom_line() +
  labs(y = NULL, x = "State") +
  theme(legend.position = "top", legend.direction = "horizontal")

plot_ic_short

# Improvement plateaus already at 3,4 states

# Decoding

# Check state interpretability and stability

## Select max likelihood model for each state choice
mod_selected_short <- comparison_table_short %>%
  filter(between(n, 2, 4))

response_table_short <- mod_selected_short %>%
  rowwise() %>%
  reframe(n = n, state_clas = 1:n, var = "y", as_tibble(matrix(getpars(mod, which = "pars")[-(1:((n + 1)*n))], ncol = 2, byrow = TRUE, dimnames = list(rep(NA, n), c("intercept", "sd"))))) %>%
  mutate(ord = rank(intercept), .by = n)

decoding_table_short <- mod_selected_short %>%
  reframe(posterior(mod[[1]], type = "viterbi"), t = data_hmm_short$time_index, y = data_hmm_short$y, .by = c(iter, n)) %>%
  mutate(state = response_table_short[response_table_short$n == cur_group()$n,]$ord[state], .by = n)

plot_comp_short <- decoding_table_short %>%
  select(n, t, y, state) %>%
  mutate(state_clas = state) %>%
  pivot_longer(c(y), names_to = "var", values_to = "val") %>%
  ggplot(aes(x = t, y = val)) +
  # geom_line(alpha = 0.3) +
  geom_point(aes(color = factor(state_clas)), alpha = 0.55, size = 1.3, pch=16) +
  geom_hline(data = response_table_short, aes(yintercept = intercept, group = factor(ord)), linewidth = 0.83) +
  geom_hline(data = response_table_short, aes(yintercept = intercept, color = factor(ord)), linewidth = 0.8) +
  facet_wrap(~ n, labeller = as_labeller(\(n) paste(n, "states"))) +
  guides(color=guide_legend(override.aes=list(size=5), title = "State")) +
  scale_color_viridis_d() +
  labs(y = "Temperature anomalies (°C)", x = NULL)

plot_comp_short

# Consequence of only modelling a level: states are splitting up the increasing trend

decoding_table_short %>%
  count(n, state, name = "count") %>%
  mutate(prop = count / sum(count) * 100, .by = n) %>%
  ggplot(aes(x = state, y = prop)) +
  geom_col() +
  facet_wrap(~ n)

walk(mod_selected_short$mod, summary)
```

```{r multi-sample trend, results = FALSE, fig.show = "hide"}
# Fit
# comparison_table_trend <- fit_models(data_hmm, max_states = 8, reps = 30, trend = TRUE)
# saveRDS(comparison_table_trend, "work/models_trend.RDS")
comparison_table_trend <- readRDS("work/models_trend.RDS")

plot_ic_trend <- comparison_table_trend %>%
  pivot_longer(c(mod_logLik, mod_BIC, mod_AIC), names_to = "indicator", values_to = "val", names_prefix = "mod_") %>%
  ggplot(aes(x = n, y = val, color = indicator)) +
  geom_line() +
  labs(y = NULL, x = "State") +
  theme(legend.position = "top", legend.direction = "horizontal")

plot_ic_trend

# BIC bottoms out at 5

# Decoding

# Check state interpretability and stability

## Select max likelihood model for each state choice
mod_selected_trend <- comparison_table_trend %>%
  filter(between(n, 3, 5))

response_table_trend <- mod_selected_trend %>%
  rowwise() %>%
  reframe(n = n, state_clas = 1:n, var = "y", as_tibble(matrix(getpars(mod, which = "pars")[-(1:((n + 1)*n))], ncol = 3, byrow = TRUE, dimnames = list(rep(NA, n), c("intercept", "slope", "sd"))))) %>%
  mutate(ord = rank(slope), .by = n)

decoding_table_trend <- mod_selected_trend %>%
  reframe(posterior(mod[[1]], type = "viterbi"), t = data_hmm$time_index, y = data_hmm$y, .by = c(iter, n)) %>%
  mutate(state = response_table_trend[response_table_trend$n == cur_group()$n,]$ord[state], .by = n)

plot_comp_trend <- decoding_table_trend %>%
  select(n, t, y, state) %>%
  mutate(state_clas = state) %>%
  pivot_longer(c(y), names_to = "var", values_to = "val") %>%
  ggplot(aes(x = t, y = val)) +
  # geom_line(alpha = 0.3) +
  geom_point(aes(color = factor(state_clas)), alpha = 0.55, size = 1.3, pch=16) +
  geom_abline(data = response_table_trend, aes(intercept = intercept, slope = slope, group = factor(ord)), linewidth = 0.83) +
  geom_abline(data = response_table_trend, aes(intercept = intercept, slope = slope, color = factor(ord)), linewidth = 0.8) +
  facet_wrap(~ n, labeller = as_labeller(\(n) paste(n, "states"))) +
  guides(color=guide_legend(override.aes=list(size=5), title = "State")) +
  scale_color_viridis_d() +
  labs(y = "Temperature anomalies (°C)", x = NULL)

plot_comp_trend

decoding_table_trend %>%
  count(n, state, name = "count") %>%
  mutate(prop = count / sum(count) * 100, .by = n) %>%
  ggplot(aes(x = state, y = prop)) +
  geom_col() +
  facet_wrap(~ n)

walk(mod_selected_trend$mod, summary)
```

```{r ic plots, out.width = "33%", fig.width = 3, fig.height = 3, fig.cap = "HMM specification information criteria by state.", fig.subcap = c("State-dependent level", "State-dependent level (short dataset)", "State-dependent linear trend")}
plot_ic_base
plot_ic_short
plot_ic_trend
```

```{r comparison plot base, out.width = "100%", fig.width = 8, fig.height = 3, fig.cap = "Decoded states for state-dependent level HMM by number of states. Lines indicate estimated state means."}
plot_comp_base
```

```{r comparison plot short, out.width = "100%", fig.width = 8, fig.height = 3, fig.cap = "Decoded states for state-dependent level HMM (short dataset) by number of states. Lines indicate estimated state means."}
plot_comp_short
```

```{r comparison plot trend, out.width = "100%", fig.width = 8, fig.height = 3, fig.cap = "Decoded states for state-dependent linear trend HMM by number of states. Lines indicate estimated state trend lines."}
plot_comp_trend
```


<!-- ```{r multi-sample detrend, output = FALSE}
data_hmm_stl <- stl(ts(data_hmm$y, frequency = 12), s.window = "periodic")

plot(data_hmm_stl$time.series[, "trend"])

data_hmm_detrend <- data_hmm %>%
  arrange(time_index) %>%
  mutate(y_det = y - data_hmm_stl$time.series[, "trend"])

data_hmm_detrend %>%
  ggplot(aes(x = time_index, y = y_det)) +
  geom_point() +
  geom_hline(yintercept = 0)

# Fit
# comparison_table_detrend <- fit_models(data_hmm_detrend, fit_var = "y_det", max_states = 8, reps = 3, trend = FALSE)
# saveRDS(comparison_table_detrend, "work/models_detrend.RDS")
comparison_table_detrend <- readRDS("work/models_detrend.RDS")

comparison_table_detrend %>%
  pivot_longer(c(mod_logLik, mod_BIC, mod_AIC), names_to = "indicator", values_to = "val", names_prefix = "mod_") %>%
  ggplot(aes(x = n, y = val)) +
  geom_line() +
  facet_wrap(~ indicator)

# BIC bottoms out at 5

# Decoding

# Check state interpretability and stability

## Select max likelihood model for each state choice
mod_selected_detrend <- comparison_table_detrend %>%
  filter(between(n, 2, 6))

decoding_table_detrend <- mod_selected_detrend %>%
  reframe(posterior(mod[[1]], type = "viterbi"), t = data_hmm_detrend$time_index, y = data_hmm_detrend$y, y_det = data_hmm_detrend$y_det, .by = n)

response_table_detrend <- mod_selected_detrend %>%
  rowwise() %>%
  reframe(n = n, state_clas = 1:n, var = "y_det", as_tibble(matrix(getpars(mod, which = "pars")[-(1:((n + 1)*n))], ncol = 2, byrow = TRUE, dimnames = list(rep(NA, n), c("intercept", "sd")))))

decoding_table_detrend %>%
  select(n, t, y, y_det, state) %>%
  mutate(state_clas = state) %>%
  pivot_longer(c(y, y_det, state), names_to = "var", values_to = "val") %>%
  ggplot(aes(x = t, y = val)) +
  geom_line(alpha = 0.3) +
  geom_point(aes(color = factor(state_clas)), alpha = 0.7, size = 1.5) +
  facet_grid(rows = vars(var), cols = vars(n), scales = "free_y") +
  guides(color=guide_legend(override.aes=list(size=8), title = "State")) +
  scale_color_viridis_d() +
  labs(y = NULL) +
  geom_hline(data = response_table_detrend, aes(yintercept = intercept, color = factor(state_clas)))

decoding_table_detrend %>%
  select(n, t, y, state) %>%
  mutate(state_clas = state) %>%
  pivot_longer(c(y, state), names_to = "var", values_to = "val") %>%
  ggplot(aes(x = t, y = val)) +
  geom_line(alpha = 0.3) +
  geom_point(aes(color = factor(state_clas)), alpha = 0.7, size = 1.5) +
  facet_grid(rows = vars(var), cols = vars(n), scales = "free_y") +
  guides(color=guide_legend(override.aes=list(size=8), title = "State")) +
  scale_color_viridis_d() +
  labs(y = NULL) +
  geom_hline(data = response_table, aes(yintercept = intercept, color = factor(state_clas)))

decoding_table_detrend %>%
  count(n, state, name = "count") %>%
  mutate(prop = count / sum(count) * 100, .by = n) %>%
  ggplot(aes(x = state, y = prop)) +
  geom_col() +
  facet_wrap(~ n)

walk(mod_selected_detrend$mod, summary)
``` -->
