---
title: "Bocconi University - 20236 Time Series Analysis"
subtitle: "**Group 27 | Anna-Bella Dreyfus, Emanuele Artoni, Artūrs Jānis Ņikitins,
  Madelief van Weerdenburg**"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
    extra_dependencies:
        - float
        - subfig
fontsize: 10pt
geometry: margin=2cm
header-includes:
- \setlength{\textfloatsep}{6pt}
- \setlength{\intextsep}{6pt}
---

```{r setup, include=FALSE}
rm(list = ls())
library(tidyverse)
library(dlm)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.pos = "H", fig.align = "center", out.extra = "", fig.width = 5, fig.height = 3, out.width = "67%")

set_theme(theme_classic())
update_theme(plot.title = element_text(hjust = 0.5), strip.background = element_blank())
```

```{r plot, fig.cap = "Annual flow of the Nile river.\\label{fig:data}"}
data_nile <- Nile

df_nile <- tibble(x = time(Nile), y = Nile)

df_nile %>%
    ggplot(aes(x, y)) + geom_line() + geom_point() + labs(x = NULL, y = "Annual flow")
```

As shown in Figure \ref{fig:data}, the series clearly appears non-stationary, presenting a quite evident change point.
A local level model, i.e. a random walk plus noise, may be used to capture the main change point and other minor changes in the level of the Nile river.

```{r setup dlm}
mod <- dlmModPoly(order=1, dV = 15100, dW = 1470, m0 = 1000, C0 = 1000)
```

# FILTERING

```{r filtering, fig.cap = "Filtered state estimate standard deviation.\\label{fig:sds}"}
mod_filt <- dlmFilter(data_nile, mod)

data_mod <- tibble(
    t = time(data_nile),
    y = data_nile,
    filt_estim = mod_filt$m[-1],
    filt_sd = sqrt(unlist(dlmSvd2var(u = mod_filt$U.C, d = mod_filt$D.C)))[-1],
    filt_ci_bot = qnorm(0.025, filt_estim, filt_sd),
    filt_ci_top = qnorm(0.975, filt_estim, filt_sd)
)

# C_t plot
data_mod %>%
    ggplot(aes(x = t, y = filt_sd)) +
    geom_line() +
    labs(x = NULL, y = "Filtered state estimate standard deviation")
```

```{r filtering-estims, fig.cap = "Annual Nile flow and filtered state.\\label{fig:filt}"}
# Data & filtering estimates & 95% CI
data_mod %>%
    ggplot(aes(x = t)) +
    geom_line(aes(y = y, color = "Data")) +
    geom_line(aes(y = filt_estim, color = "Filtered state")) +
    geom_ribbon(aes(ymin = filt_ci_bot, ymax = filt_ci_top, fill = "95% CI"), alpha = 0.2) +
    scale_color_manual(values = c(Data = "black", `Filtered state` = "red")) +
    labs(x = NULL, y = "Annual flow")
```

Figure \ref{fig:sds} shows that filtered state estimate standard deviations converge to a stable non-zero value (63.5). This is because in a random walk plus noise model, new uncertainty is added at each filtering step, so it never goes away. Figure \ref{fig:filt} shows the filtered state estimates and their 95% CIs.

# ONLINE FORECASTING

```{r forecasts, fig.cap = "Annual Nile flow and one-step-ahead forecasts.\\label{fig:frcst}"}
data_frcst <- tibble(
    t = time(data_nile),
    y = data_nile,
    frcst_estim = mod_filt$f,
    frcst_sd = sqrt(unlist(dlmSvd2var(u = mod_filt$U.R, d = mod_filt$D.R)) + mod$V),
    frcst_ci_bot = qnorm(0.025, frcst_estim, frcst_sd),
    frcst_ci_top = qnorm(0.975, frcst_estim, frcst_sd)
)

data_frcst %>%
    ggplot(aes(x = t)) +
    geom_line(aes(y = y, color = "Data")) +
    geom_line(aes(y = frcst_estim, color = "1-ahead forecast")) +
    geom_ribbon(aes(ymin = frcst_ci_bot, ymax = frcst_ci_top, fill = "95% CI"), alpha = 0.2) +
    scale_color_manual(values = c(Data = "black", `1-ahead forecast` = "red")) +
    labs(x = NULL, y = "Annual flow")
```

Figure \ref{fig:frcst} shows the one-step-ahead forecasts and their 95% CIs. For the local level model, the point estimates are equal to the previous period's filtered state estimates.

# SIGNAL-TO-NOISE

```{r signal-to-noise, fig.ncol = 2, out.width = "50%", fig.cap = "Annual Nile flow and one-step-ahead forecasts by signal-to-noise.", fig.subcap = c("High: $V = 1510, W = 1470$\\label{fig:high}", "Low: $V = 151000, W = 1470$\\label{fig:low}")}

# default: dV = 15100, dW = 1470

# High
# dV = 1510, dW = 1470

mod_high <- dlmModPoly(order=1, dV = 1510, dW = 1470, m0 = 1000, C0 = 1000)

# FILTERING

mod_filt_high <- dlmFilter(data_nile, mod_high)

data_mod_high <- tibble(
    t = time(data_nile),
    y = data_nile,
    filt_estim = mod_filt_high$m[-1],
    filt_sd = sqrt(unlist(dlmSvd2var(u = mod_filt_high$U.C, d = mod_filt_high$D.C)))[-1],
    filt_ci_bot = qnorm(0.025, filt_estim, filt_sd),
    filt_ci_top = qnorm(0.975, filt_estim, filt_sd)
)

# C_t plot
# data_mod_high %>%
    # ggplot(aes(x = t, y = filt_sd)) + geom_line()

# Data & filtering estimates & 95% CI
# data_mod_high %>%
#     ggplot(aes(x = t)) +
#     geom_line(aes(y = y, color = "Data")) +
#     geom_line(aes(y = filt_estim, color = "Filtered state")) +
#     geom_ribbon(aes(ymin = filt_ci_bot, ymax = filt_ci_top, fill = "95% CI"), alpha = 0.2) +
#     scale_color_manual(values = c(Data = "black", `Filtered state` = "red")) +
#     labs(x = NULL, y = "Annual flow")


# ONLINE FORECASTING

data_frcst_high <- tibble(
    t = time(data_nile),
    y = data_nile,
    frcst_estim = mod_filt_high$f,
    frcst_sd = sqrt(unlist(dlmSvd2var(u = mod_filt_high$U.R, d = mod_filt_high$D.R)) + mod_high$V),
    frcst_ci_bot = qnorm(0.025, frcst_estim, frcst_sd),
    frcst_ci_top = qnorm(0.975, frcst_estim, frcst_sd)
)

data_frcst_high %>%
    ggplot(aes(x = t)) +
    geom_line(aes(y = y, color = "Data")) +
    geom_line(aes(y = frcst_estim, color = "1-ahead forecast")) +
    geom_ribbon(aes(ymin = frcst_ci_bot, ymax = frcst_ci_top, fill = "95% CI"), alpha = 0.2) +
    scale_color_manual(values = c(Data = "black", `1-ahead forecast` = "red")) +
    labs(x = NULL, y = "Annual flow")

# Low
# dV = 151000, dW = 1470

mod_low <- dlmModPoly(order=1, dV = 151000, dW = 1470, m0 = 1000, C0 = 1000)

# FILTERING

mod_filt_low <- dlmFilter(data_nile, mod_low)

data_mod_low <- tibble(
    t = time(data_nile),
    y = data_nile,
    filt_estim = mod_filt_low$m[-1],
    filt_sd = sqrt(unlist(dlmSvd2var(u = mod_filt_low$U.C, d = mod_filt_low$D.C)))[-1],
    filt_ci_bot = qnorm(0.025, filt_estim, filt_sd),
    filt_ci_top = qnorm(0.975, filt_estim, filt_sd)
)

# C_t plot
# data_mod_low %>%
#     ggplot(aes(x = t, y = filt_sd)) + geom_line()

# Data & filtering estimates & 95% CI
# data_mod_low %>%
#     ggplot(aes(x = t)) +
#     geom_line(aes(y = y, color = "Data")) +
#     geom_line(aes(y = filt_estim, color = "Filtered state")) +
#     geom_ribbon(aes(ymin = filt_ci_bot, ymax = filt_ci_top, fill = "95% CI"), alpha = 0.2) +
#     scale_color_manual(values = c(Data = "black", `Filtered state` = "red")) +
#     labs(x = NULL, y = "Annual flow")


# ONLINE FORECASTING

data_frcst_low <- tibble(
    t = time(data_nile),
    y = data_nile,
    frcst_estim = mod_filt_low$f,
    frcst_sd = sqrt(unlist(dlmSvd2var(u = mod_filt_low$U.R, d = mod_filt_low$D.R)) + mod_low$V),
    frcst_ci_bot = qnorm(0.025, frcst_estim, frcst_sd),
    frcst_ci_top = qnorm(0.975, frcst_estim, frcst_sd)
)

data_frcst_low %>%
    ggplot(aes(x = t)) +
    geom_line(aes(y = y, color = "Data")) +
    geom_line(aes(y = frcst_estim, color = "1-ahead forecast")) +
    geom_ribbon(aes(ymin = frcst_ci_bot, ymax = frcst_ci_top, fill = "95% CI"), alpha = 0.2) +
    scale_color_manual(values = c(Data = "black", `1-ahead forecast` = "red")) +
    labs(x = NULL, y = "Annual flow")
```

Reducing V increases the signal-to-noise ratio (Figure \ref{fig:high}), so the fitted level follows the data more closely and forecast credible intervals become tighter. Increasing V decreases the signal-to-noise ratio (Figure \ref{fig:low}), so the model smooths more aggressively, reacts more slowly to changes, and forecast intervals become wider.

# SMOOTHING

```{r smoothing, fig.ncol = 2, out.width = "50%", fig.cap = "Annual Nile flow and smoothed state estimates.", fig.subcap = c("Full sample\\label{fig:smooth_full}", "$t = 26$ to $t = 30$\\label{fig:smooth_short}")}
mod_smooth <- dlmSmooth(mod_filt)

data_smooth <- tibble(
    t = time(data_nile),
    y = data_nile,
    smooth_estim = mod_smooth$s[-1],
    smooth_sd = sqrt(unlist(dlmSvd2var(u = mod_smooth$U.S, d = mod_smooth$D.S)))[-1],
    smooth_ci_bot = qnorm(0.025, smooth_estim, smooth_sd),
    smooth_ci_top = qnorm(0.975, smooth_estim, smooth_sd)
)

# data_smooth %>%
#     ggplot(aes(x = t)) +
#     geom_line(aes(y = y, color = "Data")) +
#     geom_line(aes(y = smooth_estim, color = "Smoothed state")) +
#     geom_ribbon(aes(ymin = smooth_ci_bot, ymax = smooth_ci_top, fill = "95% CI"), alpha = 0.2) +
#     scale_color_manual(values = c(Data = "black", `Smoothed state` = "red")) +
#     labs(x = NULL, y = "Annual flow") +
#     geom_vline(xintercept = 1898)


data_smooth %>%
    left_join(data_mod, by = c("t", "y")) %>%
    ggplot(aes(x = t)) +
    geom_line(aes(y = y, color = "Data")) +
    geom_line(aes(y = filt_estim, color = "Filtered state")) +
    geom_line(aes(y = smooth_estim, color = "Smoothed state")) +
    geom_ribbon(aes(ymin = smooth_ci_bot, ymax = smooth_ci_top, fill = "95% CI (smoother)"), alpha = 0.2) +
    geom_ribbon(aes(ymin = filt_ci_bot, ymax = filt_ci_top, fill = "95% CI (filter)"), alpha = 0.2) +
    scale_color_manual(values = c(Data = "black", `Filtered state` = "red", `Smoothed state` = "blue")) +
    labs(x = NULL, y = "Annual flow")


data_smooth %>%
left_join(data_mod, by = c("t", "y")) %>%
slice(26:30) %>%
ggplot(aes(x = t)) +
geom_line(aes(y = y, color = "Data")) +
geom_line(aes(y = filt_estim, color = "Filtered state")) +
geom_line(aes(y = smooth_estim, color = "Smoothed state")) +
geom_ribbon(aes(ymin = smooth_ci_bot, ymax = smooth_ci_top, fill = "95% CI (smoother)"), alpha = 0.2) +
geom_ribbon(aes(ymin = filt_ci_bot, ymax = filt_ci_top, fill = "95% CI (filter)"), alpha = 0.2) +
scale_color_manual(values = c(Data = "black", `Filtered state` = "red", `Smoothed state` = "blue")) +
labs(x = NULL, y = "Annual flow")

```

Figure \ref{fig:smooth_full} shows the smoothed state estimates and their 95% CIs.
At $t=28$ (1898), as shown in Figure \ref{fig:smooth_short}, the smoothed estimate (1000) is lower than the filtered estimate (1133) because smoothing uses the full sample, while filtering only uses information up to time $t$.
Around the changepoint, future observations contain important information on the subsequent lower level, so the smoothed estimate takes this into account.
Its interval is also narrower, since smoothing uses more information than filtering.

\newpage

# Code Appendix

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
