---
title: "Assignment 3 - Towards the final project"
subtitle: "Bocconi University - 20236 Time Series Analysis \\linebreak Group 27 | Anna-Bella Dreyfus, Emanuele Artoni, Artūrs Jānis Ņikitins,
  Madelief van Weerdenburg"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
    extra_dependencies:
        - float
        - subfig
fontsize: 10pt
geometry: margin=2cm
header-includes:
- \setlength{\textfloatsep}{6pt}
- \setlength{\intextsep}{6pt}
---

```{r setup, include=FALSE}
rm(list = ls())
library(tidyverse)
library(dlm)
library(ggplot2)
library(patchwork)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.pos = "H", fig.align = "center", out.extra = "", fig.width = 5, fig.height = 3, out.width = "67%")

set_theme(theme_classic())
update_theme(plot.title = element_text(hjust = 0.5), strip.background = element_blank())
```

```{r Implementation}
df <- read.csv("gistemp.csv", skip = 1, na.strings = "***")
monthly_matrix <- as.matrix(df[, 2:13])
monthly_vector <- as.numeric(t(monthly_matrix))

gistemp_ts_raw <- ts(monthly_vector, start = c(1880, 1), frequency = 12)

# Eliminating seasonality
stl_decomp <- stl(gistemp_ts_raw, s.window = "periodic")
gistemp_ts <- gistemp_ts_raw - stl_decomp$time.series[,"seasonal"]

```


# Dynamic linear models
Unlike the Hidden Markov Model from \textit{Assignment 2} which assumes abrupt discrete regime
switches, DLMs model the latent temperature as evolving continuously via
Gaussian innovations, a more natural assumption for a physical process
like the climate. We consider two specifications: a random walk plus noise
\[
Y_t = \mu_t + \epsilon_t, \quad \epsilon_t \sim \mathcal{N}(0, \sigma^2)
\]
\[
\mu_t = \mu_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, \tau^2),
\]
and a linear growth model
\[
Y_t = \mu_t + \epsilon_t, \quad \epsilon_t \sim \mathcal{N}(0, \sigma^2)
\]
\[
\mu_t = \mu_{t-1} + \beta_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, \tau^2)
\]
\[
\beta_t = \beta_{t-1} + \zeta_t, \quad \zeta_t \sim \mathcal{N}(0, \gamma^2).
\]
Both DLMs are time-invariant since the variance parameters are assumed constant over time. Therefore, changes in the estimated trend are driven exclusively by the latent state evolution.

```{r rw and llt, out.width='60%'}
build_rw <- function(params) {
  dlmModPoly(
    order = 1,
    dV = exp(params[1]),
    dW = exp(params[2])
  )
}

build_llt <- function(params) {
  dlmModPoly(
    order = 2,
    dV = exp(params[1]),
    dW = c(exp(params[2]), exp(params[3]))
  )
}
```

# Parameter estimation

```{r params estimation, out.width='60%'}
set.seed(88888)

# MLE - rw
mle_rw <- dlmMLE(gistemp_ts,
                  parm = c(0, 0),
                  build = build_rw,
                  hessian = TRUE
                  )
# output - rw
sigma2_rw <- exp(mle_rw$par[1])
tau2_rw <- exp(mle_rw$par[2])
se_rw <- sqrt(diag(solve(mle_rw$hessian))) * exp(mle_rw$par)

# MLE - llt
mle_llt <- dlmMLE(gistemp_ts,
       parm = c(0, 0, 0),
       build = build_llt,
       hessian = TRUE
       )

# output - llt
sigma2_llt <- exp(mle_llt$par[1])
tau2_llt <- exp(mle_llt$par[2])
gamma2_llt <- exp(mle_llt$par[3])
se_llt <- sqrt(diag(solve(mle_llt$hessian))) * exp(mle_llt$par)

mle_table <- data.frame(
  Parameter = c("$\\sigma^2_{RW}$", "$\\tau^2_{RW}$", "$\\sigma^2_{LLT}$", "$\\tau^2_{LLT}$", "$\\gamma^2$"),
  MLE       = c(sigma2_rw, tau2_rw, sigma2_llt, tau2_llt, gamma2_llt),
  SE        = c(se_rw, se_llt)
)
knitr::kable(mle_table, format = "latex", digits = 6, caption = "MLE estimates and asymptotic standard errors",   escape = FALSE)


```

Table 1 shows the parameters of interest and the associated standard errors. The estimate of $\gamma^2$ for the linear growth model and its standard error is extremely small and indistinguishable from zero. This suggests that the data provide little evidence for a time-varying growth rate, and, thus, that the LLT model actually behaves like a random walk plus noise whose state equation includes a constant drift component.

# Filtering and smoothing

```{r, fig.width=12, fig.height=9, fig.cap="Filtering and smoothing estimates for the RW and LLT models."}
# fitted models:
mod_rw  <- build_rw(mle_rw$par)
mod_llt <- build_llt(mle_llt$par)

## RW - Filtering

filter_sd <- function(filt) {
  vars <- dlmSvd2var(filt$U.C, filt$D.C)
  sqrt(sapply(vars, function(v) v[1, 1]))
}

rw_Filt <- dlmFilter(gistemp_ts, mod_rw)
m_filt <- dropFirst(rw_Filt$m)
rw_filt_sd <- filter_sd(rw_Filt)[-1]


z95 <- qnorm(0.975)
ci_upper <- m_filt + z95 * rw_filt_sd
ci_lower <- m_filt - z95 * rw_filt_sd

t_vals <- time(m_filt)

par(mfrow = c(2, 2), mar = c(2.8, 2, 1.8, 1), oma = c(0, 0, 0, 0))
plot(gistemp_ts, type = 'n', col = "grey60", ylab = "Anomaly (Celsius)",
     xlab = "Year", main = "RW – Filtering", cex.main = 0.8)

polygon(c(t_vals, rev(t_vals)),
        c(ci_upper, rev(ci_lower)),
        col = adjustcolor("orange", alpha.f = 0.7),
        border = NA)

points(
  t_vals,
  as.numeric(gistemp_ts),
  pch = 16,
  cex = 0.5,
  col = adjustcolor("grey25", alpha.f = 0.5)
)

lines(m_filt, col = "seagreen4", lwd = 2.2)


## RW - Smoothing

rw_Smooth <- dlmSmooth(gistemp_ts, mod_rw)

s_smooth <- dropFirst(rw_Smooth$s)
s_smooth <- as.numeric(s_smooth)

S_smooth <- dlmSvd2var(rw_Smooth$U.S, rw_Smooth$D.S)
S_smooth <- dropFirst(S_smooth)

smooth_sd <- sqrt(unlist(S_smooth))

ci_upper_smooth <- s_smooth + z95 * smooth_sd
ci_lower_smooth <- s_smooth - z95 * smooth_sd

t_vals <- time(gistemp_ts)[seq_along(s_smooth)]

plot(gistemp_ts, type = "n", col = "grey60", ylab = "Anomaly (Celsius)",
     xlab = "Year", main = "RW – Smoothing", cex.main = 0.8)

polygon(c(t_vals, rev(t_vals)),
        c(ci_upper_smooth, rev(ci_lower_smooth)),
        col = adjustcolor("orange", alpha.f = 0.7),
        border = NA)

points(
  t_vals,
  as.numeric(gistemp_ts),
  pch = 16,
  cex = 0.5,
  col = adjustcolor("grey25", alpha.f = 0.5)
)

lines(t_vals, s_smooth, col = "brown", lwd = 2.2)


## LLT - Filtering

llt_Filt <- dlmFilter(gistemp_ts, mod_llt)
m_llt <- dropFirst(llt_Filt$m[, 1])
m_llt <- as.numeric(m_llt)
C_llt <- dlmSvd2var(llt_Filt$U.C, llt_Filt$D.C)
C_llt <- dropFirst(C_llt)

llt_filt_sd <- sqrt(sapply(C_llt, function(x) x[1, 1]))


ci_upper_llt <- m_llt + z95 * llt_filt_sd
ci_lower_llt <- m_llt - z95 * llt_filt_sd

t_vals <- time(gistemp_ts)[seq_along(m_llt)]

plot(gistemp_ts, type = "n", col = "grey60", ylab = "Anomaly (Celsius)",
     xlab = "Year", main = "LLT – Filtering", cex.main = 0.8)

polygon(c(t_vals, rev(t_vals)),
        c(ci_upper_llt, rev(ci_lower_llt)),
        col = adjustcolor("orange", alpha.f = 0.4),
        border = NA)

points(
  t_vals,
  as.numeric(gistemp_ts),
  pch = 16,
  cex = 0.5,
  col = adjustcolor("grey25", alpha.f = 0.5)
)

lines(t_vals, m_llt,
      col = "seagreen",
      lwd = 2.2)

## LLT - Smoothing

llt_Smooth <- dlmSmooth(gistemp_ts, mod_llt)
s_smooth_llt <- dropFirst(llt_Smooth$s[, 1])
s_smooth_llt <- as.numeric(s_smooth_llt)
S_smooth_llt <- dlmSvd2var(llt_Smooth$U.S, llt_Smooth$D.S)
S_smooth_llt <- dropFirst(S_smooth_llt)
smooth_sd_llt <- sqrt(sapply(S_smooth_llt, function(x) x[1, 1]))

ci_upper_llt <- s_smooth_llt + z95 * smooth_sd_llt
ci_lower_llt <- s_smooth_llt - z95 * smooth_sd_llt

plot(gistemp_ts, type = "n", col = "grey60", ylab = "Anomaly (Celsius)",
     xlab = "Year", main = "LLT – Smoothing", cex.main = 0.8)

polygon(c(t_vals, rev(t_vals)),
        c(ci_upper_llt, rev(ci_lower_llt)),
        col = adjustcolor("orange", alpha.f = 0.7),
        border = NA)

points(
  t_vals,
  as.numeric(gistemp_ts),
  pch = 16,
  cex = 0.5,
  col = adjustcolor("grey25", alpha.f = 0.5)
)

lines(t_vals, s_smooth_llt, col = "brown", lwd = 2.2)
par(mfrow = c(1, 1))
```


Figure 1 shows the filtering and smoothing estimates for both models. In both the random walk plus noise and local linear trend model, it is visible that the 95% credible interval band is tighter for smoothing estimates. In the graphs, the intervals are given by the orange band around their respective green and brown (filtered and smoothed) estimates. The gray points in the graphs are the observed data. The credible intervals are tighter for smoothing rather than filtering because they are being determined by more data: in filtering, at time t, only data from time 1 through t is used, while in smoothing the entire data set from 1 through T is considered. Using 'future' information reduces uncertainty, which leads to a tighter credible interval.

# Forecasting, model checking and model evaluation

```{r forecasting, fig.width=9, fig.height=5, fig.cap="ACFs and QQ plots for the RW and LLT models." }

# Split dataset
max_date <- max(time(gistemp_ts))
training_date <- max_date - 20

gistemp_ts_train <- window(gistemp_ts, end = training_date)
gistemp_ts_test <- window(gistemp_ts, start = training_date + 1/frequency(gistemp_ts))

# Train models

set.seed(88888)

# Redo builders with log-variances

build_rw_train <- function(params) {
  dlmModPoly(order = 1, dV = exp(params[1]), dW = exp(params[2]))
}
build_llt_train <- function(params) {
  dlmModPoly(order = 2, dV = exp(params[1]), dW = c(exp(params[2]), exp(params[3])))
}

# ML estimation - rw
mle_rw_train <- dlmMLE(gistemp_ts_train,
                  parm = c(0, 0),
                  build = build_rw_train,
                  hessian = TRUE
                  )
# output - rw
sigma2_rw_train <- exp(mle_rw_train$par[1])
tau2_rw_train <- exp(mle_rw_train$par[2])

# - llt
mle_llt_train <- dlmMLE(gistemp_ts_train,
       parm = c(0, 0, 0),
       build = build_llt_train,
       hessian = TRUE
       )

# output - llt
sigma2_llt_train <- exp(mle_llt_train$par[1])
tau2_llt_train <- exp(mle_llt_train$par[2])
gamma2_llt_train <- exp(mle_llt_train$par[3])

# filters & forecast errors

mod_rw_train  <- dlmFilter(gistemp_ts, build_rw_train(mle_rw_train$par))
mod_llt_train <- dlmFilter(gistemp_ts, build_llt_train(mle_llt_train$par))

gistemp_frcst_rw <- gistemp_ts_test - window(mod_rw_train$f, start = training_date + 1/frequency(gistemp_ts))
gistemp_frcst_rw_std <- window(residuals(mod_rw_train, sd = FALSE), start = training_date + 1/frequency(gistemp_ts))

gistemp_frcst_llt <- gistemp_ts_test - window(mod_llt_train$f, start = training_date + 1/frequency(gistemp_ts))
gistemp_frcst_llt_std <- window(residuals(mod_llt_train, sd = FALSE), start = training_date + 1/frequency(gistemp_ts))

# Model checking
par(mfrow = c(2, 2), mar = c(2.8, 2, 1.8, 1), oma = c(0, 0, 0, 0))

acf(gistemp_frcst_rw_std)
acf(gistemp_frcst_llt_std)

qqnorm(gistemp_frcst_rw_std)
qqline(gistemp_frcst_rw_std)

qqnorm(gistemp_frcst_llt_std)
qqline(gistemp_frcst_llt_std)
par(mfrow = c(1, 1))

# RMSE
rmse_rw <- sqrt(mean(gistemp_frcst_rw^2))
rmse_llt <- sqrt(mean(gistemp_frcst_llt^2))

rmse_table <- data.frame(
  Model = c("RW", "LLT"),
  RMSE = c(rmse_rw, rmse_llt)
)

```

The training-sample estimates are very close to the full-sample estimates, with $\gamma^2$ still close to zero.
Figure 2 shows the ACFs and QQ plots for both models. The ACFs of the standardized forecast errors show some residual autocorrelation, with a few lags exceeding the approximate confidence bounds. This suggests that neither model fully removes temporal dependence in the test-set forecast errors. The QQ plots are reasonably close to linear in the middle part, but deviate in the tails. Thus, the normality assumption is reasonable but imperfect, especially for more extreme errors. Both models have effectively the same test-set RMSE (LLT is negligibly smaller than RW). Moreover, the variance for the estimated slope innovation is essentially zero, which suggests that the data do not support a time-varying slope component. Therefore, the LLT model does not provide a meaningful improvement over the more parsimonious RW model.

```{r}
knitr::kable(rmse_table, format = "latex", digits = 6, caption = "Test-set RMSE", escape = FALSE) |>
  kableExtra::kable_styling(latex_options = "hold_position")
```

# Discussion
State-space models assume that the series $Y_t$ is an imperfect measurement of a latent Markov chain $\theta_t$, called the state process. Conditionally on $\theta_t$, the observations are independent, and $Y_t$ depends on $\theta_t$ only. In this application, the latent process represents the underlying temperature trend.

Hidden Markov Models are state-space models in which the states are discrete-valued random variables. In contrast, dynamic linear models assume that the latent temperature trend evolves continuously and is normally distributed.

HMMs are useful for representing regime-switching behavior. For example, we used an HMM in \textit{Assignment 2} to model the nominal interest rate for 10Y Italian bonds as moving over time from a few persistent states interpreted as a recession, a boom and stable regimes. For a physical process such as the temperature trend, however, the HMM merely provides a simplified description of the series' evolution. DLMs, which are designed to recover a smoothly, continuously evolving latent trend, may be better suited to represent the underlying trend.

The empirical results support this interpretation. Both the RW and LLT models recover a clear long-term increase in temperature anomalies. The LLT model does not substantially improve on the RW model: the estimated slope innovation variance $\gamma^2$ is essentially zero, and the two models have almost identical test-set RMSEs. This suggests that the data do not require a freely time-varying growth rate, so the more parsimonious RW model is sufficient.

In terms of interpretability, a HMM's coarser representation nevertheless offers a more straightforward and synthetic representation of the latent state's evolution. Evidently, this relies heavily on whether the analyst is confident in the latent process' construct validity and in the meaning of the small number of states that they have identified. For example, identifying three economic states, from recession to booms, as strong determinants of interest rate evolution through time does make for a plain, but direct representation. DLMs are less categorical but more directly interpretable for temperature data: the latent state is the underlying temperature trend itself, and the uncertainty bands quantify how precisely this trend is estimated. Overall, the DLM framework appears more appropriate for capturing long-term temperature evolution, while the HMM framework is more useful when the goal is to identify discrete regimes.
