---
title: "Bocconi University - 20236 Time Series Analysis"
subtitle: "**Group 27 | Anna-Bella Dreyfus, Emanuele Artoni, Artūrs Jānis Ņikitins, Madelief van Weerdenburg**"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: yes
    extra_dependencies: ["float"]
fontsize: 10pt
geometry: margin=2cm
header-includes:
  - \setlength{\textfloatsep}{6pt}
  - \setlength{\intextsep}{6pt}
---

```{r setup, include=FALSE}
rm(list = ls())
library(tidyverse)
library(gridExtra)
library(depmixS4)
library(stargazer)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center", fig.pos = "H", out.extra = "")

set_theme(theme_classic())
update_theme(plot.title = element_text(hjust = 0.5), strip.background = element_blank())

```

## Task (a)

```{r, fig.height=3, fig.width=6, fig.show = "hold", out.width = "60%", fig.cap = "\\label{fig:main}nominal and real interest rates for 10Y Italian bonds"}
# extract
df <- read.csv("data_assHMM.csv", header = TRUE, sep = ",")

#turn into ts
nom_bond <- as.ts(df$ITAbond_nom)
real_bond <- as.ts(df$ITAbond_real)

# obviously we need a time index that references the actual dataset
Time_useful <- as.Date(df$Time, format = "%Y_%m_%d")

#plot stuff with ggplot
obj1_nom <- tibble(
  Time = Time_useful,
  Value = nom_bond
)
obj2_real <- tibble(
  Time = Time_useful,
  Value = real_bond
)

ggplot(obj1_nom, aes(x = Time, y = Value, color = "Nominal rate"))+
  geom_line()+
  labs(title = NULL, x= NULL, y= "interest rates for Italian 10Y bonds")+
  geom_line(data = obj2_real, aes(x = Time, y = Value, color = "Real rate"))+
  theme(legend.position = "inside", legend.position.inside = c(0.3, 0.9)) +
  labs(color = NULL)


```
We have plotted the nominal and real interest rate for Italian 10Y bonds. Since the two time series
represent a 1-1 function of each other, it is reasonable to focus our analysis exclusively on one of time,the nominal rate.
 Figure$~$\ref{fig:main} shows that, evidently, a time series that represents nominal rates for public bonds over decades cannot be described as stationary, and is also difficult to transform into a stationary one.
Inflation shocks, fiscal crises, credibility concerns, credit downgrades, unexpected shocks, political decisions all play a role in determining the sequence that cannot be simplified by first-differencing or other methods.
Thus, it is reasonable to assume that the nominal interest rate's level at time t is dependent on the unobserved state of a random process that satisfies the Markov property. In this simplified version, we then assume the existence of three hidden states: a recession, characterized by high  rates, a boom, characterized by less risky, lower interest rates, and a third state representing a stable path.

```{r fig.height=3, fig.width=6, fig.show = "hold", out.width = "60%", fig.cap = "\\label{fig:main}decoded optimal state sequence and comparison with data"}

# step 1 - model specification
# using depmix function per lab2, with three states per instructions
y <- as.numeric(nom_bond)
model <- depmix(y ~ 1, data = data.frame(y),nstates = 3)
#model

# step 2: model estimation
# using fit function from same package
fmodel <- fit(model)
#fmodel
#summary(fmodel)

# printing the 3 coefficients. note that depmix returns an object with a number of slots
# of these, we are interested in 'response', a list of states
# [[i]] selects the i-th state (there's two, the second is sd)
# [[1]] selects the first response within that state (think of row and column)
# @parameters$coefficients accesses the actual fitted coefficients
x=c(1,2,3)
for (i in x) {
  #print(c("MLE",i))
  #print(fmodel@response[[i]][[1]]@parameters$coefficients)
}

# the problem is hunting for the right parameters. there are 18 parameters in the 
# standard error output, 11 whose se is specified (free)
# 3 are initial state parameters 
# 9 are transition probs (eg from s1 to s1, from s1 to s2, from s1 to s3, times 3 equals 9)
# 3*2=6 are the response parameters (3 intercepts plus 3 sd)
# total is 18
# then print V and note that pars 13, 15 and 17 are s1_mean, s2_mean, s3_mean.
# these 20 minutes of philosophizing have led me to the following 2 loc's.
V <- standardError(fmodel)
coef_se <- V$se[c(13, 15, 17)]
#coef_se





# step 3: decoding
estStates <- posterior(fmodel)

i= estStates[1,1]
ii= if(i==1){i+1} else {i-1}
estMean1=fmodel@response[[i]][[1]]@parameters$coefficients
estMean2=fmodel@response[[ii]][[1]]@parameters$coefficients

estMeans=rep(estMean1, length(Nile))
estMeans[estStates[,1]==ii]=estMean2

# plot
# plot(time(nom_bond), estStates[,1], cex=.3)
results <- tibble(
  Time = Time_useful,
  sample_trajectory = y,
  estimated_state = estStates[,1]
  )%>%
  gather("variable","Value", -Time)

ggplot(results, aes(x = Time, y = Value))+
  geom_line()+
  facet_wrap(variable ~ ., scales="free", ncol=1)+
  labs(color = NULL)

```
Model estimation gives us the following ML estimates for each state: the coefficient associated to the "boom" state is estimated to be 2.137; the coefficient associated to the "stable" state is estimated to be 4.251; the coefficient associated to the "recession" state is estimated to be 5.640. The associated standard errors are 0.074, 0.0376, 0.0923 respectively.

The decoded sequence indicates a mix of long regimes and some periods of more frequent switching, which is consistent with a Markov regime model and with the strong persistence estimated in the transition matrix. The high-rate regime is concentrated in earlier years and around periods of stress, while the later sample is mostly assigned to lower-rate regimes.



## Task (b)




































