---
title: "Bocconi University - 20236 Time Series Analysis"
subtitle: "**Take home assignment 1 | Group 27**"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["wrapfig"]
fontsize: 10pt
geometry: margin=2cm
---

```{r, setup, include=FALSE}
library(tidyverse)
library(xts)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE)

set_theme(theme_classic())
update_theme(plot.title = element_text(hjust = 0.5), strip.background = element_blank())

# plotting functions based on ggplot
plot_ts_gg <- function(
  x,
  title = NULL,
  facet_ncol = 1,
  wrap_width = 45
) {
  if (!inherits(x, "ts")) {
    stop("Input must be ts.")
  }

  df <- tibble(
      Time = as.numeric(time(x)),
      Value = as.numeric(x)
  )

  p <- ggplot(df, aes(Time, Value)) +
    geom_line() +
    labs(title = title, x = NULL, y = NULL)

  p
}

plot_ts_dec_gg <- function(
  x,
  title = NULL,
  facet_ncol = 1,
  wrap_width = 45
) {
  if (inherits(x, "decomposed.ts")) {
    df <- tibble(
      Time = as.numeric(time(x$x)),
      Observed = as.numeric(x$x),
      Trend = as.numeric(x$trend),
      Seasonal = as.numeric(x$seasonal),
      Random = as.numeric(x$random)
    )
  } else if (inherits(x, "stl")) {
    df <- data.frame(
      Time = as.numeric(time(x$time.series)),
      Observed = as.numeric(rowSums(x$time.series)),
      Trend = as.numeric(x$time.series[, "trend"]),
      Seasonal = as.numeric(x$time.series[, "seasonal"]),
      Random = as.numeric(x$time.series[, "remainder"])
    )
  } else {
    stop("Input must be decomposed.ts or stl.")
  }

  df_long <- pivot_longer(
    df,
    cols = -Time,
    names_to = "Component",
    values_to = "Value"
  ) %>%
    mutate(Component = factor(Component, levels = c("Observed", "Trend", "Seasonal", "Random")))

  p <- ggplot(df_long, aes(Time, Value)) +
    geom_line() +
    facet_wrap(~Component, ncol = facet_ncol, scales = "free_y") +
    labs(title = title, x = NULL, y = NULL)

  p
}


plot_multi_ts_gg <- function(
  x,
  title = NULL,
  wrap_width = 45
) {
  if (!inherits(x, "ts")) {
    stop("x must be a ts or multivariate ts (e.g., cbind(...))")
  }

  df <- data.frame(Time = as.numeric(time(x)), as.data.frame(x))

  df_long <- pivot_longer(
    df,
    cols = -Time,
    names_to = "Series",
    values_to = "Value"
  )

  if (!is.null(title)) {
    title <- paste(strwrap(title, width = wrap_width), collapse = "\n")
  }

  ggplot(df_long, aes(Time, Value)) +
    geom_line() +
    facet_wrap(~Series, ncol = 1, scales = "free_y") +
    labs(title = title, x = NULL, y = NULL)
}


```

# EXERCISE 1

## EXERCISE 1.1

The following printout shows the generated time series:

```{r, exercise 1-1}
# Exercise 1.1
x <- rnorm(24, sd = sqrt(2))
y <- ts(x)
y

```

## EXERCISE 1.2

A plot's render size can be reduced with the chunk options `fig.height`, `fig.width` and display size with `out.height`, `out.width`.

## EXERCISE 1.3

STL stands for Season and Trend decomposition using LOESS. Given a univariate time series, it identifies seasonal, trend, and remainder components, assuming an additive decomposition.
LOESS estimates a smooth curve by fitting local regressions around each point, weighting nearby observations more heavily than distant ones.
The algorithm proceeds iteratively through inner and outer loops. In the inner loop, the seasonal and trend components are alternately updated using LOESS smoothing. In the outer loop, robustness weights are recomputed to reduce the influence of outliers; these updated weights are then used in the next inner-loop iteration.
STL first applies LOESS smoothing to detrend the series; then, it applies the same method separately to each seasonal subseries (e.g. to all quarters). The remainder is thus deprived of any structure.
STL differs from classical time series decomposition methods, which typically involve moving averages to estimate trends and assume seasonal patterns to be constant, in that it allows for flexibility in specifying the amounts of variation in the trend and seasonal components.

## EXERCISE 1.4

```{r, transformation, fig.width=5, fig.height=3, fig.show = "hold", out.width = "50%"}
# Exercise 1.4
log_AirPassengers <- log(AirPassengers)
comb <- cbind(log_AirPassengers, AirPassengers)

df <- tibble(Raw = AirPassengers) %>%
  mutate(
    Time = time(AirPassengers),
    Log = log(AirPassengers)
  ) %>% pivot_longer(
    c(Log, Raw),
    names_to = "Series",
    values_to = "Value"
  ) %>%
  mutate(Series = factor(Series, levels = c("Raw", "Log")))

ggplot(df, aes(Time, Value)) +
  geom_line() +
  facet_wrap(~Series, ncol = 1, scales = "free_y") +
  labs(
    title = "Raw time series (left) and log-transformed series (right)",
    x = NULL, y = NULL
  )

# Additive decomposition
m <- decompose(log_AirPassengers, type = "additive")

plot_ts_dec_gg(
  m,
  title = "Additive decomposition of log_AirPassengers",
  facet_ncol = 2
)

```

\begin{wrapfigure}[11]{r}{0.5\textwidth}

 \vspace{-\baselineskip}

 \centering

```{r, stl decomposition, fig.width=5, fig.height=3, out.width = "100%", fig.align = "right"}
s <- stl(log_AirPassengers, s.window = "periodic")

plot_ts_dec_gg(
  s,
  title = "STL decomposition of log_AirPassengers",
  facet_ncol = 2
)

```

\end{wrapfigure}

By comparing the dataset to its log-transformation, it appears obvious that the series is heteroskedastic, violating stationarity. We can then conveniently model the series by taking the log-transform and then fitting an additive decomposition. This yields the result included in the following plot.

It is also possible to take logs and apply an additive decomposition using STL. In particular, with `s.window = "periodic"`, the method forces constant seasonal patterns, making it similar to classical decomposition.

# EXERCISE 2

The following figures contain the classical decomposition of the time series and a comparison of the de-trended and stationary series, and of the residuals.

```{r, co2 decomposition, fig.width=5, fig.height=3, fig.show = "hold", out.width = "50%"}
# Exercise 2

# decompose and plot the components
m <- decompose(co2, type = "additive")

plot_ts_dec_gg(
  m,
  title = "Additive decomposition of CO2",
  facet_ncol = 2
)

df <- tibble(
  Time = as.numeric(time(co2)),
  Original = as.numeric(co2),
  Detrended = as.numeric(co2 - m$trend),
  Y_star = as.numeric(co2 - m$trend - m$seasonal),
  Residuals = as.numeric(m$random),
  Diff = Residuals - Y_star
) %>%  pivot_longer(
  cols = c(Original, Detrended, Y_star, Residuals),
  names_to = "Panel",
  values_to = "Value"
) %>%
  mutate(Panel = factor(Panel, levels = c("Original", "Detrended", "Y_star", "Residuals")))

# data for the red line
df_diff <- df %>%
  filter(Panel == "Residuals")

ggplot(df, aes(Time, Value)) +
  geom_line() +
  geom_line(data = df_diff, aes(Time, Diff), color = "red") +
  facet_wrap(~Panel, ncol = 2, scales = "free_y") +
  labs(
    title = strwrap("CO2: original, detrended, Y*, residuals (diff with Y^* in red)"),
    x = NULL,
    y = NULL
  )


```

# EXERCISE 3

The time series is clearly heteroskedastic. It is possible to either use a multiplicative decomposition, or to take the log-transform of the time series to stabilize the variance, and then apply an additive decomposition. In the results, we show the latter.

```{r, UKgas decomposition, fig.height=3, fig.width=5, fig.show = "hold", out.width = "50%"}
# Exercise 3
#1
plot_ts_gg(
  UKgas,
  title = "Raw time series"
)

#2
m <- decompose(log(UKgas), type = "additive")
plot_ts_dec_gg(
  m,
  title = "Additive decomposition of log_UKgas time series",
  facet_ncol = 2
)
```

```{r UKgas STL, fig.width=5, fig.height=3, fig.show = "hold", out.width = "50%"}
s <- stl(log(UKgas), s.window = 7)
plot_ts_dec_gg(
  s,
  title = "STL (additive) decomposition of UKgas",
  facet_ncol = 2
)

## HW decomposition
hw <- HoltWinters(log(UKgas),seasonal = "additive")

df_hw <- tibble(
  Time = as.numeric(time(UKgas)),
  Original = hw$x,
  Fitted = c(rep(NA, 4), hw$fitted[, "xhat"])
) %>%
  pivot_longer(-Time, names_to = "var", values_to = "val")

ggplot(df_hw, aes(x = Time, y = val, color = var)) +
  geom_line() +
  labs(
    title = "Holt-Winters decomposition",
    x = NULL,
    y = NULL
  ) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.1, 0.9),
    legend.title = element_blank()
  ) +
  scale_color_manual(values = c("Original" = "black", "Fitted" = "red"))

```

The classical additive decomposition shows that the residuals' distribution over time is non-random around 1970, suggesting that the model's assumption of constant seasonal amplitude is violated. In particular, since classical decomposition constrains the seasonal component to be time-invariant, any change in seasonal pattern is absorbed by the residuals. To confirm and address this issue, we can run an alternative STL decomposition to the same log-transformed series, a method which allows the seasonal component to vary over time. We can thus clearly determine that the seasonal amplitude indeed changes over time, and that the rate of change also increases significantly after the early 1970's. This is an indication that the classical time series decomposition is inadequate at identifying the evolving seasonal pattern.

```{r UKgas HW, include=FALSE}
hw <- HoltWinters(log(UKgas), seasonal = "additive")
note <- sprintf(
  "alpha = %.3f, beta = %.3f, gamma = %.3f",
  hw$alpha,
  hw$beta,
  hw$gamma
)

```

After applying an additive model to the log-transformation of the series, the fitted Holt-Winters procedure tracks both the trend and the seasonal variation, indicating that a level-trend-seasonality exponential smoothing procedure fits the data well. In particular, the decomposition fixes alpha = `r sprintf("%.3f", hw$alpha)`, beta = `r sprintf("%.3f", hw$beta)`, gamma = `r sprintf("%.3f", hw$gamma)`.

# EXERCISE 4

```{r, simple ES, fig.width=5, fig.height=3, fig.show = "hold", out.width = "50%"}
# Exercise 4
m <- HoltWinters(Nile, beta = FALSE, gamma = FALSE)
s <- HoltWinters(Nile, alpha = 0.1, beta = FALSE, gamma = FALSE)
t <- HoltWinters(Nile, alpha = 0.9, beta = FALSE, gamma = FALSE)

mape = function(y, yhat) {
  mean(abs(y - yhat) / y)
}

y_m <- window(Nile, start = time(Nile)[2])
m_mape <- mape(y_m, m$fitted[, 1])
s_mape <- mape(y_m, s$fitted[, 1])
t_mape <- mape(y_m, t$fitted[, 1])

# plot stuff
df <- tibble(
  Time = as.numeric(time(Nile)),
  Original = as.numeric(Nile),
  `Default α` = c(NA, as.numeric(m$fitted[, "xhat"])),
  `α = 0.1` = c(NA, as.numeric(s$fitted[, "xhat"])),
  `α = 0.9` = c(NA, as.numeric(t$fitted[, "xhat"]))
) %>%
  pivot_longer(-Time, names_to = "var", values_to = "val")

df %>%
  filter(var %in% c("Original", "Default α"))  %>%
  ggplot(aes(x = Time, y = val, color = var)) +
    geom_line() +
    scale_color_manual(
      values = c("Original" = "black", "Default α" = "red")
    ) +
    labs(
      title = "Simple Exponential Smoothing (Nile)",
      x = NULL,
      y = NULL,
      color = NULL,
      linetype = NULL
    ) +
    theme(legend.position = "inside", legend.position.inside = c(0.125, 0.2))

df %>%
  filter(var %in% c("Original", "α = 0.1", "α = 0.9"))  %>%
  ggplot(aes(x = Time, y = val, color = var, linetype = var)) +
    geom_line() +
    scale_color_manual(
      values = c("Original" = "black", "α = 0.1" = "red", "α = 0.9" = "blue")
    ) +
    scale_linetype_manual(
      values = c("Original" = "solid", "α = 0.1" = "dashed", "α = 0.9" = "dashed")
    ) +
    labs(
      title = "Comparison of smoothing strength",
      x = NULL,
      y = NULL,
      color = NULL,
      linetype = NULL
    ) +
    theme(legend.position = "inside", legend.position.inside = c(0.1, 0.2))

```

For alpha = 0.2466, the MAPE is 0.1307; for alpha = 0.1, the MAPE is 0.1348; for alpha = 0.9, the MAPE is 0.1450.
Using Holt-Winters with beta = gamma = 0, the estimated alpha is 0.2466. As the smoothing parameter, alpha controls the responsiveness of the level estimate. When alpha = 0.1, the prediction depends on many earlier values, making for stable and slow predictions. When alpha = 0.9, the model is highly adaptive, reacting strongly to the most recent observation. Thus, forecasts adjust quickly, but are also susceptible to noise. Therefore, it makes sense that the value chosen by the model to minimize the loss function is a compromise between the two. In fact, the drop observed at the end of the recorded series is captured most strongly by setting alpha = 0.9, and least so by setting alpha = 0.1, which yields the highest predicted value.
In particular, the MAPE is lowest for alpha = 0.2466; highest for alpha = 0.9, confirming that the model's chosen alpha maximizes predictive performance.

# EXERCISE 5

```{r, COVID data, fig.width=10, fig.height=3}
# Exercise 5
# extraction, cleaning
data_covid <- read.table("coronavirus-data.txt", header = TRUE, fill = TRUE)

dates <- seq.Date(
  as.Date("2020-02-20"),
  by = "day",
  length.out = nrow(data_covid)
)
xts_data_covid <- xts(data_covid |> dplyr::select(-day), order.by = dates)

# plot for all vars
df_wide <- xts_data_covid |> fortify.zoo() # transform xts into dataframe, as ggplot works with latter only
df_plot <- df_wide |>
  pivot_longer(-Index, names_to = "variable", values_to = "value")

ggplot(df_plot, aes(x = Index, y = value)) +
  geom_line() +
  facet_wrap(~variable, scales = "free_y") +
  labs(x = NULL, y = NULL)

# new vars of interest, non-cumulative
new_cases <- diff(data_covid$contagion, lag = 1)
new_deaths <- diff(data_covid$deaths, lag = 1)

# plot for new vars
df_new <- data.frame(
  date = dates[-1],
  new_cases = new_cases,
  new_deaths = new_deaths
)


df_long <- pivot_longer(
  df_new,
  cols = c(new_cases, new_deaths),
  names_to = "Panel",
  values_to = "value"
)

ggplot(df_long, aes(x = date, y = value)) +
  geom_line() +
  facet_wrap(~Panel, ncol = 2, scales = "free_y") +
  labs(
    title = "Daily new cases and deaths",
    x = NULL,
    y = NULL
  )

```

```{r COVID decom, fig.width=10, fig.height=4}
# decomposition analysis + plot
cases_ts <- ts(new_cases, start = c(2020, 8), frequency = 7)
deaths_ts <- ts(new_deaths, start = c(2020, 8), frequency = 7)
stl_fit <- stl(cases_ts, s.window = 7)

stl_ts <- cbind(
  Observed = cases_ts,
  Trend = stl_fit$time.series[, "trend"],
  Seasonal = stl_fit$time.series[, "seasonal"],
  Remainder = stl_fit$time.series[, "remainder"]
)
plot_multi_ts_gg(
  stl_ts,
  title = "STL decomposition of daily new cases"
)

```

```{r COVID ses}
# 1st ES for the whole sample
ses_fit <- HoltWinters(cases_ts, beta = TRUE, gamma = TRUE)
#ses_fit$alpha

# 2nd ES for sub-sample from 08-01 before fall wave
start_date <- as.Date("2020-08-01")

idx <- dates[-1] >= start_date

new_cases_aug <- new_cases[idx]
dates_aug <- dates[-1][idx]

cases_ts_aug <- ts(new_cases_aug, frequency = 7)

ses_fit2 <- HoltWinters(cases_ts_aug, beta = TRUE, gamma = TRUE)
#ses_fit2

```

The raw data shows cumulative cases (contagion), deaths, and strong growth in testing over time. It is also clear that intensive care facilities experienced epidemic waves.
Since the contagion and deaths measures cumulative flows, we have constructed a differenced time series of daily new cases.
Graphically, this series displays clear epidemic waves, in particular in the later months of 2020. STL decomposition confirms the presence of a trend component and a clear weekly seasonal component, which we believe to be related to reporting effects.
Simple exponential smoothing via Holt-Winters yields a smoothing parameter alpha = 0.88, implying weak smoothing. Restricting the sample to observations from 1 August 2020 reduces alpha to 0.41, implying that the epidemic wave that occurred in the latter part of that year is associated with much stronger and smoother trend and seasonal components.
Overall, all the series are non-stationary and characterized by wavy effects.

# Appendix: All code for this report

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
