---
title: "Bocconi University - 20236 Time Series Analysis"
subtitle: "**Group 27 | Anna-Bella Dreyfus, Emanuele Artoni, Artūrs Jānis Ņikitins,
  Madelief van Weerdenburg**"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
    extra_dependencies: float
  html_document:
    df_print: paged
fontsize: 10pt
geometry: margin=2cm
header-includes:
- \setlength{\textfloatsep}{6pt}
- \setlength{\intextsep}{6pt}
---

```{r setup, include=FALSE}
rm(list = ls())
library(tidyverse)
library(gridExtra)
library(depmixS4)
library(stargazer)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center", fig.pos = "H", out.extra = "")

set_theme(theme_classic())
update_theme(plot.title = element_text(hjust = 0.5), strip.background = element_blank())

```

## Task (a)

```{r, fig.height=3, fig.width=6, fig.show = "hold", out.width = "60%", fig.cap = "\\label{fig:main}nominal and real interest rates for 10Y Italian bonds"}
# extract
df <- read.csv("data_assHMM.csv", header = TRUE, sep = ",")

#turn into ts
nom_bond <- as.ts(df$ITAbond_nom)
real_bond <- as.ts(df$ITAbond_real)

# obviously we need a time index that references the actual dataset
Time_useful <- as.Date(df$Time, format = "%Y_%m_%d")

#plot stuff with ggplot
obj1_nom <- tibble(
  Time = Time_useful,
  Value = nom_bond
)
obj2_real <- tibble(
  Time = Time_useful,
  Value = real_bond
)

ggplot(obj1_nom, aes(x = Time, y = Value, color = "Nominal rate"))+
  geom_line()+
  labs(title = NULL, x= NULL, y= "interest rates for Italian 10Y bonds")+
  geom_line(data = obj2_real, aes(x = Time, y = Value, color = "Real rate"))+
  scale_color_manual(values = c(
    "Nominal rate" = "black",
    "Real rate" = "blue"
  )) +
  theme(legend.position = "inside", legend.position.inside = c(0.3, 0.9)) +
  labs(color = NULL)


```
We have plotted the nominal and real interest rate for Italian 10Y bonds. Since the two time series
are closely related, it is reasonable to focus our analysis exclusively focus our analysis on one of them, namely the nominal rate.
Figure$~$\ref{fig:main} shows that, evidently, a time series that represents nominal rates for public bonds over decades cannot be described as stationary, and is also difficult to transform into a stationary one.
Inflation shocks, fiscal crises, credibility concerns, credit downgrades, unexpected shocks, political decisions all play a role in determining the sequence that cannot be simplified by first-differencing or other methods.
Thus, it is reasonable to assume that the nominal interest rate's level at time t is dependent on the unobserved state of a random process that satisfies the Markov property. In this simplified version, we then assume the existence of three hidden states: a recession, characterized by high  rates, a boom, characterized by less risky, lower interest rates, and a third state representing a stable path.

```{r fig.height=3, fig.width=6, fig.show = "hold", out.width = "60%"}

# step 1 - model specification
# using depmix function per lab2, with three states per instructions
y <- as.numeric(nom_bond)
model <- depmix(y ~ 1, data = data.frame(y),nstates = 3)
#model

```

```{r, include=FALSE}
set.seed(555)
fmodel <- fit(model)
#summary(fmodel)
```


```{r}
# printing the 3 coefficients. note that depmix returns an object with a number of slots
# of these, we are interested in 'response', a list of states
# [[i]] selects the i-th state of 3
# [[1]] selects the first response within that state (first is mean, second is sd)
# @parameters$coefficients accesses the actual fitted coefficients
ml_coeffs <- sapply(1:3, function(i) {
  fmodel@response[[i]][[1]]@parameters$coefficients
})
sds <- sapply(1:3, function(i) {
  fmodel@response[[i]][[1]]@parameters$sd
})
# the problem is hunting for the right parameters. there are 18 parameters in the 
# standard error output, 11 whose se is specified (free)
# 3 are initial state parameters 
# 9 are transition probs (eg from s1 to s1, from s1 to s2, from s1 to s3, times 3 equals 9)
# 3*2=6 are the response parameters (3 intercepts plus 3 sd)
# total is 18
# then print V and note that pars 13, 15 and 17 are s1_mean, s2_mean, s3_mean.
V <- standardError(fmodel)
coef_se <- V$se[c(13, 15, 17)]
#coef_se

# as the previous comment made clear, the transition probs are included in the 
# standarderror output. in particular, if the first three pars are the initial
# probabilities, then we need the 9 that come after (4:12)
transition_prob <- matrix(V$par[4:12], nrow = 3, ncol = 3, byrow = TRUE)
rownames(transition_prob) <- paste0("fromS", 1:3)
colnames(transition_prob) <- paste0("toS", 1:3)

transition_prob

table_out <- data.frame(
  State = paste("State", 1:3),
  Mean = round(ml_coeffs, 3),
  SE_Mean = round(coef_se, 3),
  SD = round(sds, 3)
)
table_out

```

After specifying the model, we have fitted it by the expectation maximization algorithm. In total, there are 18 parameters: 3 initial probabilities, 9 transition probabilities and 6 response parameters, i.e. the intercepts and the associated standard deviations.

The first table presents the transition probabilities. It indicates strong persistence, particularly for states 2 and 3, as reflected in the high diagonal probabilities.
The second table presents the estimates for each state and its associated standard deviation and SE measures.

State 1 can be interpreted as the recession regime, as it presents the highest mean rate, and the highest standard error too, indicating more volatility and risk.
State 3 may refer to the stable regime, as it has the lowest standard deviation; state 2 is instead characterized by the lowest mean nominal rate, allowing us to interpret it as the boomimg regime.

```{r fig.height=3, fig.width=6, fig.show = "hold", out.width = "60%", fig.cap = "\\label{fig:second}decoded optimal state sequence and comparison with data"}

# step 3: decoding
estStates <- posterior(fmodel, type = "viterbi")

estMean1_rec=fmodel@response[[1]][[1]]@parameters$coefficients
estMean2_boom=fmodel@response[[2]][[1]]@parameters$coefficients
estMean3_stable=fmodel@response[[3]][[1]]@parameters$coefficients

estMeans=rep(estMean1_rec, length(nom_bond))
estMeans[estStates[,1]==2]=estMean2_boom
estMeans[estStates[,1]==3]=estMean3_stable



# plot
# plot(time(nom_bond), estStates[,1], cex=.3)
results <- tibble(
  Time = Time_useful,
  sample_trajectory = nom_bond,
  estimated_state_mean = estMeans
  )%>%
  gather("variable","Value", -Time)

ggplot(results, aes(x = Time, y = Value))+
  geom_line()+
  facet_wrap(variable ~ ., scales="free", ncol=1)+
  labs(color = NULL)
```
In Figure$~$\ref{fig:second}, We have plotted the trajectory of nominal bond rates. Along with it, since the state indexing (1-3) has no economic interpretation, we have plotted the trajectory of the estimated mean associated with each decoded hidden state using the Viterbi algorithm. The decoded sequence indicates a mix of long regimes and some periods of more frequent switching, which is consistent with a Markov regime model and with the strong persistence estimated in the transition matrix.
The model identifies distinct regimes corresponding to low, intermediate, and high interest rate levels, and the timing of these regimes broadly aligns with the observed movements in the data, capturing periods of higher rates as well as more stable and lower-rate environments.
